---
title: "Using eDNAModel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using eDNAModel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(eDNAModel)
```

## Introduction

The `eDNAModel` package provides a pipeline to analyze environmental DNA (eDNA) data for species occupancy and abundance modeling using TMB.

This vignette demonstrates:

1. Loading a phyloseq object
2. Converting to a 3D array
3. Filtering the array
4. Running the TMB pipeline

---

## Load a phyloseq object

We'll use an example `.RDS` file located in `inst/extdata/`.

```{r load-data, eval = FALSE}
phyloseq_path <- system.file("extdata", "longdataexample.RDS", package = "eDNAModel")
data_array <- data_array_phyloseq(phyloseq_path)
```

---

## Filter the 3D Data Array

We remove species with all-zero counts, sites with all-zero counts, and rare species.

```{r filter-data, eval = FALSE}
filtered_array <- filter_data_array(data_array, min_species_sum = 30)
```

---

## Run the Full TMB Pipeline

Fit the statistical model to the filtered array:

```{r run-tmb, eval = FALSE}
result <- run_full_TMB_pipeline(filtered_array)
```

You can extract model output such as slopes, intercepts, and model fit metrics from `result`.

---

## Summary

This vignette showed how to:

- Convert phyloseq data to 3D array
- Filter the array
- Run TMB model pipeline

Explore the `eDNAModel` documentation for further customization and model diagnostics.
