---
title: "Occupancy, Detection, and Multivariate Community Modelling with eDNAModel"
author: "Marwah Soliman, Bert van der Veen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Occupancy, Detection, and Multivariate Community Modelling with eDNAModel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE
)
```

```{r load-packages, message=FALSE, warning=FALSE}
library(phyloseq)
library(Matrix)
library(eDNAModel)
library(dplyr)
library(tidyr)
library(ggplot2)
library(pheatmap)
library(tibble)
library(gllvm)
library(glmmTMB)
library(reshape2)
library(grid)
```

# 1. Introduction

Environmental DNA (eDNA) metabarcoding data provide an increasingly important
source of information for biodiversity monitoring. However, interpretation of
eDNA observations is complicated by imperfect detection, variation in
sequencing abundance, differences among taxa, and multivariate dependence among
species.

An observed zero does not necessarily imply that a taxon is absent from the
sampled environment. A species may be present but fail to be detected, or it
may be detected inconsistently because of biological, environmental, sampling,
or laboratory processes.

The `eDNAModel` package provides tools for modelling these different components
of eDNA data.

This vignette demonstrates the workflow implemented in `FitModel_gllvm()`.
The function combines occupancy and detection modelling with a Generalized
Linear Latent Variable Model (GLLVM) representation of multivariate community
structure.

The workflow demonstrated here includes:

1. loading a `phyloseq` object;
2. fitting the model using `FitModel_gllvm()`;
3. obtaining repeated model fits;
4. discarding initial iterations using a burn-in period;
5. estimating OTU-specific occupancy probabilities;
6. estimating OTU-specific detection probabilities;
7. constructing uncertainty intervals;
8. visualizing occupancy probabilities using caterpillar plots;
9. visualizing detection probabilities using caterpillar plots; and
10. examining multivariate community structure using GLLVM latent variables.

The example uses the **Marine Invasive species Trondheim** dataset distributed
with the `eDNAModel` package.

# 2. Conceptual Model

The modelling framework separates several processes that can contribute to
variation in eDNA observations.

For taxon \(k\), site or sampling unit \(i\), and observation \(j\), the
framework distinguishes:

- occupancy;
- detection or capture; and
- observed abundance.

This distinction is important because absence, non-detection, and low abundance
are not necessarily equivalent ecological processes.

## 2.1 Occupancy

Let

\[
Z_{ik} \in \{0,1\}
\]

denote the latent occupancy state of OTU \(k\) at site \(i\).

The occupancy state can be represented as

\[
Z_{ik} \sim \mathrm{Bernoulli}(\psi_{ik}),
\]

where

\[
\psi_{ik} = P(Z_{ik}=1)
\]

is the probability that OTU \(k\) occupies site \(i\).

The occupancy component can depend on environmental, temporal, spatial, or
sampling covariates.

For example, an occupancy linear predictor can be written as

\[
\mathrm{logit}(\psi_{ik})
=
\eta^{(\psi)}_{ik}.
\]

The corresponding occupancy probability is

\[
\psi_{ik}
=
\frac{\exp(\eta^{(\psi)}_{ik})}
     {1+\exp(\eta^{(\psi)}_{ik})}.
\]

In R, this transformation is performed using `plogis()`.

## 2.2 Detection or capture

Conditional on occurrence, a taxon may or may not be detected.

Let

\[
D_{ik} \in \{0,1\}
\]

represent detection of taxon \(k\).

The detection probability can be denoted by

\[
p_{ik}=P(D_{ik}=1 \mid Z_{ik}=1).
\]

In the output used in this vignette, the detection linear predictor is
transformed using a complementary log-log inverse link:

\[
p_{ik}
=
1-\exp\{-\exp(\eta^{(p)}_{ik})\}.
\]

Therefore, the corresponding R transformation is

```r
1 - exp(-exp(eta))
```

rather than `plogis(eta)`.

This distinction is important when transforming model predictions to the
probability scale.

## 2.3 Abundance

Conditional abundance represents variation in observed taxon counts.

A generic abundance model can be represented as

\[
Y_{ik} \sim F(\mu_{ik}),
\]

where \(F\) denotes the selected count distribution.

For a log-linked model,

\[
\log(\mu_{ik})
=
\eta^{(\mu)}_{ik}.
\]

The current example uses

```r
abundance_family = "poisson"
```

so the conditional abundance component is fitted using a Poisson distribution.

# 3. Why Use a GLLVM?

Community datasets contain many taxa measured simultaneously. Species may
respond similarly to environmental conditions or may exhibit residual
associations that are not explained by measured predictors.

Generalized Linear Latent Variable Models provide a low-dimensional
representation of this multivariate structure.

Instead of modelling every possible pairwise association directly, a small
number of latent variables can be introduced.

Conceptually, the linear predictor for taxon \(k\) at site \(i\) can contain

\[
\eta_{ik}
=
\mathbf{x}_{i}^{T}\boldsymbol{\beta}_{k}
+
\mathbf{u}_{i}^{T}\boldsymbol{\lambda}_{k},
\]

where:

- \(\mathbf{x}_{i}\) contains observed predictors;
- \(\boldsymbol{\beta}_{k}\) contains taxon-specific coefficients;
- \(\mathbf{u}_{i}\) contains latent-variable scores for sites or samples;
- \(\boldsymbol{\lambda}_{k}\) contains taxon loadings.

The latent variables summarize residual multivariate structure that remains
after accounting for measured predictors.

With two latent variables, the fitted site and species scores can be displayed
in a two-dimensional biplot.

# 4. Load Required Packages

The main packages used in this vignette are:

```{r packages}
library(eDNAModel)
library(phyloseq)
library(dplyr)
library(ggplot2)
```

The example dataset is supplied directly with `eDNAModel`, so the vignette does
not depend on a local file path.

# 5. Load and Subset the Example Data

Load the example data using:

```{r load-data}
data("physeq_new", package = "eDNAModel")

physeq_one <- physeq_new$`Marine Invasive species Trondheim`
```

The resulting object is a `phyloseq` object containing the OTU abundance table
and associated sample information.

The object can be inspected using:

```{r inspect-data}
physeq_one

phyloseq::ntaxa(physeq_one)

phyloseq::nsamples(physeq_one)

phyloseq::sample_variables(physeq_one)
```

The first rows of the sample metadata can be examined with:

```{r inspect-metadata}
head(
  as.data.frame(
    phyloseq::sample_data(physeq_one)
  )
)
```

Before fitting the model, users should verify that the variables referenced in
the model specification are present in the sample metadata.

For the current analysis, the important variables include:

- `Name`, used to identify sites or sampling units;
- `Samplingmonth`, used as an occupancy covariate and in the abundance model;
- OTU identifiers; and
- observed taxon counts.

# 6. Fit the Occupancy--Detection Model

The model is fitted using `FitModel_gllvm()`.

```{r fit-model, echo=TRUE, eval=FALSE}
out <- FitModel_gllvm(
  phyloseq = physeq_one,

  site_col = "Name",

  abundance_rhs =
    (1 | Samplingmonth / OTU) +
    (1 | OTU),

  occupancy_covars = "Samplingmonth",

  abundance_family = "poisson",

  n_iter = 50,
  burn_in = 10,
  num_lv_c = 2
)
```

For package-vignette construction, this chunk may be set to `eval=FALSE` if the
model is computationally expensive. When performing an actual analysis, run
the code normally.

# 7. Understanding the Model Specification

The arguments supplied to `FitModel_gllvm()` determine the structure of the
model.

## 7.1 `phyloseq`

```r
phyloseq = physeq_one
```

specifies the `phyloseq` object containing the community data and associated
sample metadata.

## 7.2 Site definition

```r
site_col = "Name"
```

defines the variable used as the site or sampling-unit identifier.

The scientific interpretation of the occupancy probability depends directly on
this choice.

With `Name` as the site identifier, occupancy is interpreted relative to the
sampling units represented by `Name`.

## 7.3 Abundance model

The abundance model is specified as

```r
abundance_rhs =
  (1 | Samplingmonth / OTU) +
  (1 | OTU)
```

This structure allows the abundance component to represent variation associated
with sampling month and OTUs.

The nested notation

```r
Samplingmonth / OTU
```

represents hierarchical variation involving OTUs within sampling month, while

```r
(1 | OTU)
```

allows additional OTU-level heterogeneity.

The precise interpretation of these random effects should always be considered
relative to the sampling design and the implementation of the model.

## 7.4 Occupancy covariate

The model uses

```r
occupancy_covars = "Samplingmonth"
```

so sampling month is included as a predictor of occupancy.

This allows estimated occurrence probabilities to vary according to temporal
sampling conditions represented by `Samplingmonth`.

## 7.5 Abundance family

The abundance family is

```r
abundance_family = "poisson"
```

which assumes a Poisson conditional count model.

If substantial conditional overdispersion remains, an alternative count
distribution may be appropriate when supported by the implementation and
scientific context.

## 7.6 Number of iterations

The model uses

```r
n_iter = 50
```

which requests 50 repeated iterations of the modelling procedure.

Repeated fitting allows uncertainty and variability in the fitted quantities to
be summarized across retained iterations.

## 7.7 Burn-in

The argument

```r
burn_in = 10
```

indicates that the initial iterations are excluded from the summaries produced
from retained fits.

With

```r
n_iter = 50
burn_in = 10
```

the intended summaries are based on the post-burn-in successful iterations.

It is important to distinguish this repeated-fitting procedure from an MCMC
posterior sampler unless the underlying implementation explicitly defines it as
such. Consequently, the empirical quantiles calculated below describe
variation across the retained model iterations.

## 7.8 Number of latent variables

The argument

```r
num_lv_c = 2
```

requests two latent-variable dimensions.

This is particularly useful for visualization because the first two latent
dimensions can be displayed directly in a two-dimensional biplot.

# 8. Inspecting the Model Output

After fitting, inspect the returned object before constructing figures.

```{r inspect-output, eval=FALSE}
class(out)

names(out)
```

Several components are particularly useful for the visualizations in this
vignette:

```{r important-output, eval=FALSE}
out$psi_list

out$p_detect_list

out$mean_lv_sites

out$mean_lv_species
```

Conceptually:

- `psi_list` contains retained occupancy linear predictors;
- `p_detect_list` contains retained detection linear predictors;
- `mean_lv_sites` contains averaged latent-variable coordinates for sites or
  samples;
- `mean_lv_species` contains averaged latent-variable coordinates or loadings
  for taxa.

Users should inspect the structure of these components before assuming that
specific identifier columns are available.

For example:

```{r inspect-output-structure, eval=FALSE}
str(out$psi_list, max.level = 1)

str(out$p_detect_list, max.level = 1)

head(out$mean_lv_sites)

head(out$mean_lv_species)
```

# 9. Occupancy Probability

The occupancy output is stored on the linear-predictor scale.

The first step is therefore to combine the retained iterations.

```{r occupancy-data}
otu_col <- "OTU"

psi_draws <- dplyr::bind_rows(
  out$psi_list
)
```

The occupancy predictor uses the logit scale, so it is transformed using

\[
\psi
=
\frac{\exp(\eta)}
     {1+\exp(\eta)}
=
\mathrm{logit}^{-1}(\eta).
\]

In R:

```{r occupancy-transform}
psi_draws <- psi_draws %>%
  dplyr::mutate(
    psi = stats::plogis(.data$eta)
  )
```

The resulting `psi` variable lies between 0 and 1.

Values close to zero indicate low estimated occupancy probability, whereas
values close to one indicate high estimated occupancy probability.

# 10. Summarizing Occupancy by OTU

Occupancy estimates can be summarized across sites and retained iterations for
each OTU.

```{r occupancy-summary}
psi_otu_summary <- psi_draws %>%
  dplyr::group_by(.data[[otu_col]]) %>%
  dplyr::summarise(
    psi_mean = mean(
      .data$psi,
      na.rm = TRUE
    ),

    psi_median = stats::median(
      .data$psi,
      na.rm = TRUE
    ),

    psi_lwr = stats::quantile(
      .data$psi,
      0.025,
      na.rm = TRUE,
      names = FALSE
    ),

    psi_upr = stats::quantile(
      .data$psi,
      0.975,
      na.rm = TRUE,
      names = FALSE
    ),

    .groups = "drop"
  ) %>%
  dplyr::arrange(.data$psi_mean)
```

For every OTU, this produces:

- `psi_mean`: mean estimated occupancy probability;
- `psi_median`: median estimated occupancy probability;
- `psi_lwr`: 2.5th percentile;
- `psi_upr`: 97.5th percentile.

These lower and upper quantiles describe the empirical spread of the retained
occupancy predictions represented in `psi_list`.

They should therefore not automatically be described as Bayesian posterior
credible intervals unless the fitted object is generated by a posterior
sampling procedure with that interpretation.

The results can be inspected using:

```{r inspect-occupancy-summary}
head(psi_otu_summary)

summary(psi_otu_summary$psi_mean)
```

# 11. Occupancy Probability Caterpillar Plot

OTUs are ordered from the lowest to highest estimated mean occupancy.

```{r order-occupancy}
psi_otu_summary[[otu_col]] <- factor(
  psi_otu_summary[[otu_col]],
  levels = psi_otu_summary[[otu_col]]
)
```

The caterpillar plot is then constructed as follows.

```{r plot-occupancy, fig.width=10, fig.height=9}
ggplot(
  psi_otu_summary,
  aes(
    x = psi_mean,
    y = .data[[otu_col]]
  )
) +
  geom_point(
    color = "darkgreen",
    size = 2
  ) +
  geom_errorbarh(
    aes(
      xmin = psi_lwr,
      xmax = psi_upr
    ),
    height = 0.2,
    color = "darkgreen"
  ) +
  labs(
    title = "Caterpillar Plot: Occupancy Probability by OTU",
    x = expression(
      "Occupancy probability (" * psi * ")"
    ),
    y = "OTU"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    plot.title = element_text(hjust = 0.5)
  )
```

## 11.1 Interpreting the occupancy plot

Each point represents the mean estimated occupancy probability for one OTU.

The horizontal interval represents the 2.5th to 97.5th percentile range across
the retained occupancy predictions.

OTUs toward the right side of the figure have higher estimated occupancy
probabilities.

OTUs toward the left have lower estimated occupancy probabilities.

Wide intervals indicate greater variation or uncertainty in the retained
predictions, whereas narrow intervals indicate more stable estimates.

Importantly, low occupancy should not automatically be interpreted as low
detection. Occupancy and detection describe different components of the
observation process.

# 12. Detection Probability

The detection predictions are stored in

```r
out$p_detect_list
```

The retained detection predictors are first combined.

```{r detection-data}
p_detect_draws <- dplyr::bind_rows(
  out$p_detect_list
)
```

In this output, `eta` is on the complementary log-log scale.

Therefore,

\[
\log[-\log(1-p)]
=
\eta.
\]

Solving for \(p\) gives

\[
p
=
1-\exp[-\exp(\eta)].
\]

The transformation is implemented as:

```{r detection-transform}
p_detect_draws <- p_detect_draws %>%
  dplyr::mutate(
    p_detect =
      1 - exp(-exp(.data$eta))
  )
```

The resulting detection probabilities lie between 0 and 1.

# 13. Summarizing Detection Probability by OTU

Detection probabilities are summarized in the same way as occupancy
probabilities.

```{r detection-summary}
p_detect_otu_summary <- p_detect_draws %>%
  dplyr::group_by(.data[[otu_col]]) %>%
  dplyr::summarise(
    p_detect_mean = mean(
      .data$p_detect,
      na.rm = TRUE
    ),

    p_detect_median = stats::median(
      .data$p_detect,
      na.rm = TRUE
    ),

    p_detect_lwr = stats::quantile(
      .data$p_detect,
      0.025,
      na.rm = TRUE,
      names = FALSE
    ),

    p_detect_upr = stats::quantile(
      .data$p_detect,
      0.975,
      na.rm = TRUE,
      names = FALSE
    ),

    .groups = "drop"
  ) %>%
  dplyr::arrange(.data$p_detect_mean)
```

The resulting table contains taxon-specific estimates of:

- mean detection probability;
- median detection probability;
- lower empirical quantile; and
- upper empirical quantile.

Inspect the results using:

```{r inspect-detection-summary}
head(p_detect_otu_summary)

summary(
  p_detect_otu_summary$p_detect_mean
)
```

# 14. Detection Probability Caterpillar Plot

Order the OTUs according to their estimated mean detection probability.

```{r order-detection}
p_detect_otu_summary[[otu_col]] <- factor(
  p_detect_otu_summary[[otu_col]],
  levels = p_detect_otu_summary[[otu_col]]
)
```

Then construct the caterpillar plot.

```{r plot-detection, fig.width=10, fig.height=9}
ggplot(
  p_detect_otu_summary,
  aes(
    x = p_detect_mean,
    y = .data[[otu_col]]
  )
) +
  geom_point(
    color = "steelblue",
    size = 2
  ) +
  geom_errorbarh(
    aes(
      xmin = p_detect_lwr,
      xmax = p_detect_upr
    ),
    height = 0.2,
    color = "steelblue"
  ) +
  labs(
    title =
      "Caterpillar Plot: Detection Probability by OTU",
    x = "Detection Probability",
    y = "OTU"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    plot.title = element_text(hjust = 0.5)
  )
```

## 14.1 Interpreting the detection plot

An OTU can have high occupancy but relatively low detection probability.

Such a taxon may occur widely but be difficult to detect using the sampling or
laboratory procedure.

Conversely, an OTU can have relatively restricted occupancy but high detection
probability when present.

The two caterpillar plots should therefore be interpreted jointly.

For example:

| Occupancy | Detection | Possible interpretation |
|:---|:---|:---|
| High | High | Widespread and readily detected |
| High | Low | Widespread but difficult to detect |
| Low | High | Restricted occurrence but readily detected when present |
| Low | Low | Restricted occurrence and difficult detection |

These interpretations should be considered together with the biological and
sampling context.

# 15. GLLVM Latent Variables

The GLLVM component provides a low-dimensional representation of multivariate
community structure.

With

```r
num_lv_c = 2
```

the model estimates two latent dimensions.

The site coordinates are stored in:

```r
out$mean_lv_sites
```

and the species coordinates are stored in:

```r
out$mean_lv_species
```

The first two dimensions are denoted:

```r
LV1
LV2
```

Before constructing a biplot, verify that these columns exist.

# 16. GLLVM Latent-Variable Biplot

## 16.1 Check the latent dimensions

```{r biplot-check}
required_lv <- c(
  "LV1",
  "LV2"
)

if (
  is.null(out$mean_lv_sites) ||
  is.null(out$mean_lv_species) ||
  !all(
    required_lv %in%
      names(out$mean_lv_sites)
  ) ||
  !all(
    required_lv %in%
      names(out$mean_lv_species)
  )
) {

  stop(
    "The biplot requires LV1 and LV2 in both ",
    "out$mean_lv_sites and out$mean_lv_species. ",
    "Fit the model with num_lv_c = 2."
  )
}
```

This check prevents the plotting code from continuing when fewer than two
latent dimensions are available.

## 16.2 Extract latent coordinates

```{r biplot-coordinates}
site_mat <- as.matrix(
  out$mean_lv_sites[
    ,
    required_lv,
    drop = FALSE
  ]
)

species_mat <- as.matrix(
  out$mean_lv_species[
    ,
    required_lv,
    drop = FALSE
  ]
)
```

`site_mat` contains the coordinates for samples or sites.

`species_mat` contains the corresponding taxon coordinates.

# 17. Obtaining Robust Site and OTU Labels

The latent-variable output should not be assumed to contain the original sample
metadata columns.

For example, code such as

```r
out$mean_lv_sites$Sampling.area.Name
```

can fail when `Sampling.area.Name` is not stored in the latent-variable output.

A more robust strategy is to search for available identifier columns and then
fall back to row names or generated labels.

```{r biplot-labels}
site_id_candidates <- setdiff(
  names(out$mean_lv_sites),
  required_lv
)

species_id_candidates <- setdiff(
  names(out$mean_lv_species),
  required_lv
)

if (length(site_id_candidates) > 0) {

  site_labels <- as.character(
    out$mean_lv_sites[
      [site_id_candidates[1]]
    ]
  )

} else if (
  !is.null(
    rownames(out$mean_lv_sites)
  ) &&
  length(
    rownames(out$mean_lv_sites)
  ) == nrow(site_mat)
) {

  site_labels <-
    rownames(out$mean_lv_sites)

} else {

  site_labels <- paste0(
    "Sample_",
    seq_len(nrow(site_mat))
  )
}

if (
  "OTU" %in%
    names(out$mean_lv_species)
) {

  species_labels <- as.character(
    out$mean_lv_species$OTU
  )

} else if (
  length(species_id_candidates) > 0
) {

  species_labels <- as.character(
    out$mean_lv_species[
      [species_id_candidates[1]]
    ]
  )

} else if (
  !is.null(
    rownames(out$mean_lv_species)
  ) &&
  length(
    rownames(out$mean_lv_species)
  ) == nrow(species_mat)
) {

  species_labels <-
    rownames(out$mean_lv_species)

} else {

  species_labels <- paste0(
    "OTU_",
    seq_len(nrow(species_mat))
  )
}
```

Final safety checks ensure that the number of labels equals the number of
coordinates.

```{r label-checks}
if (
  length(site_labels) !=
    nrow(site_mat)
) {

  site_labels <- paste0(
    "Sample_",
    seq_len(nrow(site_mat))
  )
}

if (
  length(species_labels) !=
    nrow(species_mat)
) {

  species_labels <- paste0(
    "OTU_",
    seq_len(nrow(species_mat))
  )
}
```

This makes the plotting procedure more robust to changes in the structure of
the model output.

# 18. Scaling the Biplot

Site scores and taxon scores may naturally occupy different numerical ranges.

Plotting the unscaled values can therefore cause one set of coordinates to
dominate the figure.

A common symmetric rescaling is used here.

First calculate the maximum Euclidean radius of the site scores:

\[
R_{\mathrm{site}}
=
\max_i
\sqrt{
LV1_i^2 + LV2_i^2
}.
\]

Similarly, for species:

\[
R_{\mathrm{species}}
=
\max_k
\sqrt{
LV1_k^2 + LV2_k^2
}.
\]

```{r biplot-radii}
site_radius <- max(
  sqrt(
    rowSums(site_mat^2)
  ),
  na.rm = TRUE
)

species_radius <- max(
  sqrt(
    rowSums(species_mat^2)
  ),
  na.rm = TRUE
)
```

The common scaling factor is

\[
c
=
\sqrt{
\frac{
R_{\mathrm{site}}
}{
R_{\mathrm{species}}
}
}.
\]

```{r biplot-scale-factor}
if (
  is.finite(site_radius) &&
  is.finite(species_radius) &&
  site_radius > 0 &&
  species_radius > 0
) {

  scale_factor <- sqrt(
    site_radius /
      species_radius
  )

} else {

  scale_factor <- 1
}
```

Site scores and species scores are then transformed symmetrically:

\[
\mathbf{u}^{*}
=
\frac{\mathbf{u}}{c},
\]

and

\[
\boldsymbol{\lambda}^{*}
=
c\boldsymbol{\lambda}.
\]

```{r biplot-scaling}
scaled_sites <-
  site_mat / scale_factor

scaled_species <-
  species_mat * scale_factor
```

This scaling is primarily for visualization. It does not refit the model or
change the underlying fitted latent structure.

# 19. Construct Biplot Data

Create plotting data frames for sites and species.

```{r biplot-data}
scaled_sites_df <- data.frame(
  Site = site_labels,
  LV1 = scaled_sites[, 1],
  LV2 = scaled_sites[, 2],
  stringsAsFactors = FALSE
)

scaled_species_df <- data.frame(
  OTU = species_labels,
  LV1 = scaled_species[, 1],
  LV2 = scaled_species[, 2],
  stringsAsFactors = FALSE
)
```

The resulting objects can be inspected before plotting.

```{r inspect-biplot-data}
head(scaled_sites_df)

head(scaled_species_df)
```

# 20. Plot the GLLVM Latent Space

```{r biplot, fig.width=10, fig.height=9}
ggplot() +

  geom_point(
    data = scaled_sites_df,
    aes(
      x = LV1,
      y = LV2
    ),
    color = "blue",
    size = 2
  ) +

  geom_text(
    data = scaled_sites_df,
    aes(
      x = LV1,
      y = LV2,
      label = Site
    ),
    color = "blue",
    size = 2.5,
    vjust = -0.5
  ) +

  geom_segment(
    data = scaled_species_df,
    aes(
      x = 0,
      y = 0,
      xend = LV1,
      yend = LV2
    ),
    arrow = grid::arrow(
      length =
        grid::unit(
          0.2,
          "cm"
        )
    ),
    color = "brown",
    alpha = 0.8
  ) +

  geom_text(
    data = scaled_species_df,
    aes(
      x = LV1,
      y = LV2,
      label = OTU
    ),
    color = "brown",
    size = 3,
    vjust = -0.5
  ) +

  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = 0.3
  ) +

  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 0.3
  ) +

  coord_equal() +

  labs(
    x = "Latent Variable 1",
    y = "Latent Variable 2",
    title =
      "GLLVM Latent Variable Biplot"
  ) +

  theme_minimal()
```

# 21. Interpreting the GLLVM Biplot

The biplot provides a graphical representation of multivariate structure in the
community.

## 21.1 Sample points

Blue points represent samples or sites.

Samples positioned close together in latent space have relatively similar
latent community profiles.

Samples located far apart differ more strongly along the latent dimensions.

Clusters of samples can therefore indicate groups with similar residual
community composition.

## 21.2 OTU arrows

Brown arrows represent taxa.

The direction of an arrow describes the direction in latent space associated
with that taxon's score or loading.

The length of an arrow reflects the magnitude of its representation in the
displayed latent dimensions after the plotting rescaling.

Taxa with arrows pointing in similar directions have similar patterns with
respect to these latent dimensions.

Taxa pointing in opposite directions exhibit contrasting patterns along the
displayed latent space.

## 21.3 Samples and taxa together

A sample positioned in the direction of a particular OTU arrow is associated
with a relatively stronger position along the latent dimensions represented by
that taxon.

However, the biplot should not automatically be interpreted as evidence of
direct ecological interactions among taxa.

Latent associations can arise from:

- unmeasured environmental gradients;
- shared habitat preferences;
- temporal variation;
- spatial structure;
- sampling effects;
- biological associations; or
- other residual sources of community dependence.

The biplot therefore describes multivariate structure rather than establishing
causal relationships.

# 22. Relationship Between Occupancy, Detection, and Latent Structure

The three principal outputs provide complementary information.

## Occupancy

Occupancy describes whether taxa are likely to occur in the sampled units.

It addresses the question:

> How likely is this OTU to occur in the sampled system?

## Detection

Detection describes the probability of observing a taxon conditional on the
relevant occupancy/detection process.

It addresses the question:

> When an OTU is present, how readily is it detected?

## GLLVM latent variables

The latent-variable component describes multivariate community structure that
is not represented directly by simple taxon-specific occupancy probabilities.

It addresses the question:

> Which samples and taxa show similar residual multivariate patterns?

These components should therefore be interpreted together rather than as
competing summaries.

# 23. Recommended Diagnostic Checks

Before interpreting the figures, inspect the fitted object carefully.

At minimum, users should determine:

```{r diagnostic-checks, eval=FALSE}
# Available output
names(out)

# Number of retained iterations
length(out$psi_list)

length(out$p_detect_list)

# Occupancy output
head(
  dplyr::bind_rows(
    out$psi_list
  )
)

# Detection output
head(
  dplyr::bind_rows(
    out$p_detect_list
  )
)

# Latent-variable output
head(out$mean_lv_sites)

head(out$mean_lv_species)
```

It is also useful to check for non-finite predictions:

```{r finite-checks, eval=FALSE}
psi_check <-
  dplyr::bind_rows(
    out$psi_list
  )

detect_check <-
  dplyr::bind_rows(
    out$p_detect_list
  )

sum(
  !is.finite(psi_check$eta)
)

sum(
  !is.finite(detect_check$eta)
)
```

A model should not be interpreted solely because the plotting code executes
successfully.

The quality of the underlying model fits and the number of successful retained
iterations should also be evaluated.

# 24. Common Problems

## 24.1 `Samplingmonth` not found

If an error indicates that `Samplingmonth` does not exist, inspect:

```r
phyloseq::sample_variables(
  physeq_one
)
```

and verify that the metadata column is named exactly `Samplingmonth`.

R variable names are case sensitive.

## 24.2 No successful iterations remain after burn-in

If the model reports that no successful iterations remain after burn-in, the
problem is with model fitting rather than plotting.

Possible steps include:

- checking the data structure;
- checking for very rare OTUs;
- increasing the number of iterations;
- examining the abundance model;
- simplifying the random-effect structure; and
- checking whether the selected latent-variable dimension is supported by the
  available data.

## 24.3 `LV1` or `LV2` is missing

The biplot requires two latent dimensions.

Fit the model using:

```r
num_lv_c = 2
```

and confirm:

```r
names(out$mean_lv_sites)

names(out$mean_lv_species)
```

## 24.4 Site labels have length zero

Do not assume that `out$mean_lv_sites` contains a particular original metadata
column.

The robust label-selection procedure used in this vignette checks available
identifier columns, then row names, and finally constructs generic sample
labels.

# 25. Recommended Analysis Workflow

A practical workflow for `FitModel_gllvm()` is:

1. Construct and validate the `phyloseq` object.
2. Verify OTU counts and sample metadata.
3. Define the appropriate site variable.
4. Select scientifically meaningful occupancy covariates.
5. Specify the abundance model.
6. Select an appropriate abundance family.
7. Choose the number of latent variables.
8. Fit the repeated models.
9. Remove the specified burn-in iterations.
10. Verify how many successful iterations remain.
11. Transform occupancy predictors to probabilities using the correct inverse
    link.
12. Transform detection predictors using the complementary log-log inverse
    link.
13. Summarize occupancy and detection by OTU.
14. Inspect uncertainty across retained iterations.
15. Examine latent-variable site and taxon scores.
16. Construct the GLLVM biplot.
17. Interpret occupancy, detection, abundance, and latent structure jointly.
18. Report the model specification and fitting settings.

# 26. Complete Example

The complete workflow can be summarized as follows.

```{r complete-example, eval=FALSE}
library(eDNAModel)
library(phyloseq)
library(dplyr)
library(ggplot2)

# Load example data
data(
  "physeq_new",
  package = "eDNAModel"
)

physeq_one <-
  physeq_new$`Marine Invasive species Trondheim`

# Fit model
out <- FitModel_gllvm(
  phyloseq = physeq_one,

  site_col = "Name",

  abundance_rhs =
    (1 | Samplingmonth / OTU) +
    (1 | OTU),

  occupancy_covars =
    "Samplingmonth",

  abundance_family =
    "poisson",

  n_iter = 50,

  burn_in = 10,

  num_lv_c = 2
)

# ---------------------------
# Occupancy
# ---------------------------

psi_draws <-
  dplyr::bind_rows(
    out$psi_list
  ) %>%
  dplyr::mutate(
    psi =
      stats::plogis(.data$eta)
  )

psi_otu_summary <-
  psi_draws %>%
  dplyr::group_by(OTU) %>%
  dplyr::summarise(
    psi_mean =
      mean(psi, na.rm = TRUE),

    psi_median =
      median(psi, na.rm = TRUE),

    psi_lwr =
      quantile(
        psi,
        0.025,
        na.rm = TRUE,
        names = FALSE
      ),

    psi_upr =
      quantile(
        psi,
        0.975,
        na.rm = TRUE,
        names = FALSE
      ),

    .groups = "drop"
  )

# ---------------------------
# Detection
# ---------------------------

p_detect_draws <-
  dplyr::bind_rows(
    out$p_detect_list
  ) %>%
  dplyr::mutate(
    p_detect =
      1 - exp(-exp(.data$eta))
  )

p_detect_otu_summary <-
  p_detect_draws %>%
  dplyr::group_by(OTU) %>%
  dplyr::summarise(
    p_detect_mean =
      mean(
        p_detect,
        na.rm = TRUE
      ),

    p_detect_median =
      median(
        p_detect,
        na.rm = TRUE
      ),

    p_detect_lwr =
      quantile(
        p_detect,
        0.025,
        na.rm = TRUE,
        names = FALSE
      ),

    p_detect_upr =
      quantile(
        p_detect,
        0.975,
        na.rm = TRUE,
        names = FALSE
      ),

    .groups = "drop"
  )

# Inspect results
psi_otu_summary

p_detect_otu_summary

out$mean_lv_sites

out$mean_lv_species
```

# 27. Reporting Results

When reporting an analysis performed with `FitModel_gllvm()`, the following
information should be documented.

## Data structure

Report:

- the dataset;
- the number of samples;
- the number of OTUs;
- the definition of the site or sampling unit; and
- relevant repeated-sampling structure.

## Model structure

Report:

- `site_col`;
- occupancy covariates;
- abundance model;
- abundance family;
- random-effect structure; and
- number of latent variables.

For the example in this vignette:

```text
Site variable: Name
Occupancy covariate: Samplingmonth
Abundance family: Poisson
Number of latent variables: 2
Number of iterations: 50
Burn-in: 10
```

## Model-derived quantities

Report the method used to obtain:

- occupancy probabilities;
- detection probabilities;
- OTU-level summaries;
- empirical uncertainty intervals;
- site latent scores; and
- taxon latent scores.

The link functions should also be stated explicitly.

For occupancy:

\[
\psi
=
\mathrm{logit}^{-1}(\eta).
\]

For detection:

\[
p
=
1-\exp[-\exp(\eta)].
\]

# 28. Summary

`FitModel_gllvm()` provides a framework for analysing multivariate eDNA
community data while distinguishing important components of the ecological and
observation process.

In the workflow presented here, occupancy represents the probability that an
OTU occurs within a sampling unit, while detection describes the probability of
detecting an OTU through the observation process. Conditional abundance is
modelled separately using an appropriate count distribution.

The GLLVM component complements these taxon-specific quantities by providing a
low-dimensional representation of residual multivariate community structure.
With two latent variables, site scores and taxon scores can be displayed
together in a biplot, facilitating exploration of similarities among samples
and taxa.

The occupancy and detection caterpillar plots provide complementary
taxon-specific summaries. A species may have high estimated occupancy but low
detection probability, or low occupancy but high detection when present.
Separating these processes helps prevent ecological absence from being
confounded with imperfect observation.

Careful attention should be given to the definition of the sampling unit, the
choice of covariates, the abundance distribution, the number of retained model
iterations, and the interpretation of the latent-variable structure.

Together, these components provide a flexible workflow for investigating
occupancy, detectability, abundance, and multivariate community structure in
eDNA metabarcoding datasets.
