---
title: "eDNAModel: gllvm"
author: "Marwah Soliman, Bert van der Veen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{eDNAModel: gllvm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load-packages, message=FALSE, warning=FALSE}
library(phyloseq)
library(Matrix)
library(eDNAModel)
library(dplyr)
library(tidyr)
library(ggplot2)
library(pheatmap)
library(tibble)
library(gllvm)
library(glmmTMB)
library(reshape2)
library(grid)
```

# 1. Introduction

This vignette demonstrates the updated occupancy--detection model workflow
using the `eDNAModel` package and `phyloseq` data objects. The workflow
includes fitting occupancy, capture, and abundance models across repeated
iterations, visualizing occupancy and detection probabilities, and using
GLLVM latent variables to describe multivariate structure among taxa and
samples.

# 2. Load and Subset Data

The example data are supplied with the `eDNAModel` package. This makes the
vignette reproducible without relying on files stored on the user's local
computer.

```{r load-data}
data("physeq_new", package = "eDNAModel")

physeq_one <- physeq_new$`Marine Invasive species Trondheim`
```

# 3. Fit the Occupancy--Detection Model

```{r fit-model, echo=FALSE}
out <- FitModel_gllvm(
  phyloseq = physeq_one,
  site_col = "Name",

  abundance_rhs =
    (1 | Samplingmonth / OTU) +
    (1 | OTU),

  occupancy_covars = "Samplingmonth",

  abundance_family = "poisson",

  n_iter = 50,
  burn_in = 10,
  num_lv_c = 2
)
```

# 4. Visualization

## 4.1 Occupancy Probability Caterpillar Plot

```{r plot-occupancy, fig.width=10, fig.height=9}
otu_col <- "OTU"

# Combine retained GLLVM occupancy predictions
psi_draws <- dplyr::bind_rows(out$psi_list)

# eta is on the logit scale
psi_draws <- psi_draws %>%
  dplyr::mutate(
    psi = stats::plogis(.data$eta)
  )

# Summarise occupancy by OTU across sites and retained iterations
psi_otu_summary <- psi_draws %>%
  dplyr::group_by(.data[[otu_col]]) %>%
  dplyr::summarise(
    psi_mean = mean(.data$psi, na.rm = TRUE),
    psi_median = stats::median(.data$psi, na.rm = TRUE),
    psi_lwr = stats::quantile(
      .data$psi,
      0.025,
      na.rm = TRUE,
      names = FALSE
    ),
    psi_upr = stats::quantile(
      .data$psi,
      0.975,
      na.rm = TRUE,
      names = FALSE
    ),
    .groups = "drop"
  ) %>%
  dplyr::arrange(.data$psi_mean)

# Order OTUs by estimated occupancy
psi_otu_summary[[otu_col]] <- factor(
  psi_otu_summary[[otu_col]],
  levels = psi_otu_summary[[otu_col]]
)

ggplot(
  psi_otu_summary,
  aes(
    x = psi_mean,
    y = .data[[otu_col]]
  )
) +
  geom_point(
    color = "darkgreen",
    size = 2
  ) +
  geom_errorbarh(
    aes(
      xmin = psi_lwr,
      xmax = psi_upr
    ),
    height = 0.2,
    color = "darkgreen"
  ) +
  labs(
    title = "Caterpillar Plot: Occupancy Probability by OTU",
    x = expression("Occupancy probability (" * psi * ")"),
    y = "OTU"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    plot.title = element_text(hjust = 0.5)
  )
```

## 4.2 Detection Probability Caterpillar Plot

```{r plot-detection, fig.width=10, fig.height=9}
otu_col <- "OTU"

# Combine retained detection predictors
p_detect_draws <- dplyr::bind_rows(out$p_detect_list)

# eta is on the complementary log-log scale
p_detect_draws <- p_detect_draws %>%
  dplyr::mutate(
    p_detect = 1 - exp(-exp(.data$eta))
  )

# Summarise detection probability by OTU
p_detect_otu_summary <- p_detect_draws %>%
  dplyr::group_by(.data[[otu_col]]) %>%
  dplyr::summarise(
    p_detect_mean = mean(.data$p_detect, na.rm = TRUE),
    p_detect_median = stats::median(.data$p_detect, na.rm = TRUE),
    p_detect_lwr = stats::quantile(
      .data$p_detect,
      0.025,
      na.rm = TRUE,
      names = FALSE
    ),
    p_detect_upr = stats::quantile(
      .data$p_detect,
      0.975,
      na.rm = TRUE,
      names = FALSE
    ),
    .groups = "drop"
  ) %>%
  dplyr::arrange(.data$p_detect_mean)

# Order OTUs
p_detect_otu_summary[[otu_col]] <- factor(
  p_detect_otu_summary[[otu_col]],
  levels = p_detect_otu_summary[[otu_col]]
)

ggplot(
  p_detect_otu_summary,
  aes(
    x = p_detect_mean,
    y = .data[[otu_col]]
  )
) +
  geom_point(
    color = "steelblue",
    size = 2
  ) +
  geom_errorbarh(
    aes(
      xmin = p_detect_lwr,
      xmax = p_detect_upr
    ),
    height = 0.2,
    color = "steelblue"
  ) +
  labs(
    title = "Caterpillar Plot: Detection Probability by OTU",
    x = "Detection Probability",
    y = "OTU"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    plot.title = element_text(hjust = 0.5)
  )
```

## 4.3 GLLVM Latent-Variable Biplot

The following biplot displays the first two latent-variable dimensions for
samples and taxa. Sample scores are represented by points and taxon scores by
arrows.

```{r biplot, fig.width=10, fig.height=9}
# ------------------------------------------------------------
# 1. Check that two latent-variable dimensions are available
# ------------------------------------------------------------

required_lv <- c("LV1", "LV2")

if (
  is.null(out$mean_lv_sites) ||
  is.null(out$mean_lv_species) ||
  !all(required_lv %in% names(out$mean_lv_sites)) ||
  !all(required_lv %in% names(out$mean_lv_species))
) {
  stop(
    "The biplot requires LV1 and LV2 in both ",
    "out$mean_lv_sites and out$mean_lv_species. ",
    "Fit the model with num_lv_c = 2."
  )
}

# ------------------------------------------------------------
# 2. Extract latent coordinates
# ------------------------------------------------------------

site_mat <- as.matrix(
  out$mean_lv_sites[, required_lv, drop = FALSE]
)

species_mat <- as.matrix(
  out$mean_lv_species[, required_lv, drop = FALSE]
)

# ------------------------------------------------------------
# 3. Obtain labels
#
# Do not assume that mean_lv_sites contains the original
# Sampling.area.Name column. Use an available identifier column
# or row names instead.
# ------------------------------------------------------------

site_id_candidates <- setdiff(
  names(out$mean_lv_sites),
  required_lv
)

species_id_candidates <- setdiff(
  names(out$mean_lv_species),
  required_lv
)

if (length(site_id_candidates) > 0) {

  site_labels <- as.character(
    out$mean_lv_sites[[site_id_candidates[1]]]
  )

} else if (
  !is.null(rownames(out$mean_lv_sites)) &&
  length(rownames(out$mean_lv_sites)) == nrow(site_mat)
) {

  site_labels <- rownames(out$mean_lv_sites)

} else {

  site_labels <- paste0(
    "Sample_", seq_len(nrow(site_mat))
  )
}

if ("OTU" %in% names(out$mean_lv_species)) {

  species_labels <- as.character(
    out$mean_lv_species$OTU
  )

} else if (length(species_id_candidates) > 0) {

  species_labels <- as.character(
    out$mean_lv_species[[species_id_candidates[1]]]
  )

} else if (
  !is.null(rownames(out$mean_lv_species)) &&
  length(rownames(out$mean_lv_species)) == nrow(species_mat)
) {

  species_labels <- rownames(out$mean_lv_species)

} else {

  species_labels <- paste0(
    "OTU_", seq_len(nrow(species_mat))
  )
}

# Final safety checks
if (length(site_labels) != nrow(site_mat)) {
  site_labels <- paste0(
    "Sample_", seq_len(nrow(site_mat))
  )
}

if (length(species_labels) != nrow(species_mat)) {
  species_labels <- paste0(
    "OTU_", seq_len(nrow(species_mat))
  )
}

# ------------------------------------------------------------
# 4. Common symmetric scaling
# ------------------------------------------------------------

site_radius <- max(
  sqrt(rowSums(site_mat^2)),
  na.rm = TRUE
)

species_radius <- max(
  sqrt(rowSums(species_mat^2)),
  na.rm = TRUE
)

if (
  is.finite(site_radius) &&
  is.finite(species_radius) &&
  site_radius > 0 &&
  species_radius > 0
) {

  scale_factor <- sqrt(
    site_radius / species_radius
  )

} else {

  scale_factor <- 1
}

scaled_sites <- site_mat / scale_factor
scaled_species <- species_mat * scale_factor

# ------------------------------------------------------------
# 5. Construct plotting data
# ------------------------------------------------------------

scaled_sites_df <- data.frame(
  Site = site_labels,
  LV1 = scaled_sites[, 1],
  LV2 = scaled_sites[, 2],
  stringsAsFactors = FALSE
)

scaled_species_df <- data.frame(
  OTU = species_labels,
  LV1 = scaled_species[, 1],
  LV2 = scaled_species[, 2],
  stringsAsFactors = FALSE
)

# ------------------------------------------------------------
# 6. Plot
# ------------------------------------------------------------

ggplot() +

  geom_point(
    data = scaled_sites_df,
    aes(x = LV1, y = LV2),
    color = "blue",
    size = 2
  ) +

  geom_text(
    data = scaled_sites_df,
    aes(
      x = LV1,
      y = LV2,
      label = Site
    ),
    color = "blue",
    size = 2.5,
    vjust = -0.5
  ) +

  geom_segment(
    data = scaled_species_df,
    aes(
      x = 0,
      y = 0,
      xend = LV1,
      yend = LV2
    ),
    arrow = grid::arrow(
      length = grid::unit(0.2, "cm")
    ),
    color = "brown",
    alpha = 0.8
  ) +

  geom_text(
    data = scaled_species_df,
    aes(
      x = LV1,
      y = LV2,
      label = OTU
    ),
    color = "brown",
    size = 3,
    vjust = -0.5
  ) +

  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = 0.3
  ) +

  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 0.3
  ) +

  coord_equal() +

  labs(
    x = "Latent Variable 1",
    y = "Latent Variable 2",
    title = "GLLVM Latent Variable Biplot"
  ) +

  theme_minimal()
```
