---
title: "Joint Occupancy, Capture, and Abundance Models with eDNAModel"
author: "Marwah Soliman, Bert van der Veen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Joint Occupancy, Capture, and Abundance Models with eDNAModel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

Environmental DNA (eDNA) metabarcoding data are naturally hierarchical. A
species may be present at a site but fail to enter a collected biological
sample, and DNA from a captured species may subsequently generate a variable
number of sequencing reads.

Consequently, zeros in an eDNA count table can arise through several different
ecological and observation processes.

The `FitModel_joint()` function in **eDNAModel** provides a joint hierarchical
framework for modelling three components of this process:

1. **Occupancy** — whether an OTU is present at a site.
2. **Capture** — whether the OTU is represented in a biological sample,
   conditional on occupancy.
3. **Abundance** — the observed sequencing count, conditional on the
   hierarchical observation process.

The function fits these components jointly using Template Model Builder (TMB).
Latent occupancy and biological-sample capture states are analytically
marginalized, while enabled Gaussian random effects are integrated using the
Laplace approximation.


# Statistical model

Consider OTU \(k\), site \(i\), biological sample \(j\), and sequencing
observation \(r\).

The model separates site-level occurrence from biological capture and from the
conditional sequencing-count process.


## Occupancy

Let \(Z_{ik} \in \{0,1\}\) denote the latent occupancy state of OTU \(k\) at
site \(i\):

\[
Z_{ik} \sim \mathrm{Bernoulli}(\psi_{ik}).
\]

The occupancy probability is modelled as

\[
\mathrm{logit}(\psi_{ik})
=
\mathbf{x}_{ik}^{(\mathrm{occ})T}
\boldsymbol{\beta}_{\mathrm{occ}}
+
b_k^{(\mathrm{occ})}.
\]

Here:

- \(\psi_{ik}\) is the probability that OTU \(k\) occupies site \(i\);
- \(\mathbf{x}_{ik}^{(\mathrm{occ})}\) contains occupancy covariates;
- \(\boldsymbol{\beta}_{\mathrm{occ}}\) contains fixed-effect coefficients;
- \(b_k^{(\mathrm{occ})}\) is an optional OTU-level random effect.


## Capture

Conditional on occupancy, let \(A_{ijk} \in \{0,1\}\) represent whether OTU
\(k\) is captured in biological sample \(j\):

\[
A_{ijk} \mid Z_{ik}=1
\sim
\mathrm{Bernoulli}(p_{ijk}).
\]

The capture probability is modelled as

\[
\mathrm{logit}(p_{ijk})
=
\mathbf{x}_{ijk}^{(\mathrm{cap})T}
\boldsymbol{\beta}_{\mathrm{cap}}
+
b_k^{(\mathrm{cap})}.
\]

The capture component therefore separates site-level occurrence from the
probability that an occupied OTU is represented in a particular biological
sample.


## Abundance

Conditional sequencing abundance is represented through a log-linear
predictor:

\[
\log(\lambda_{rjk})
=
\mathbf{x}_{rjk}^{(\mathrm{abund})T}
\boldsymbol{\beta}_{\mathrm{abund}}
+
b_k^{(\mathrm{abund})}
+
b_j^{(\mathrm{sample})}
+
b_{jk}^{(\mathrm{sample:OTU})}
+
o_{rjk}.
\]

Here:

- \(\lambda_{rjk}\) is the conditional expected sequencing count;
- \(\mathbf{x}_{rjk}^{(\mathrm{abund})}\) contains abundance covariates;
- \(b_k^{(\mathrm{abund})}\) represents OTU-level abundance heterogeneity;
- \(b_j^{(\mathrm{sample})}\) represents biological-sample heterogeneity;
- \(b_{jk}^{(\mathrm{sample:OTU})}\) represents optional sample-by-OTU
  heterogeneity;
- \(o_{rjk}\) is an optional abundance offset.

Depending on the selected family, the conditional count distribution can be
Poisson, negative binomial, zero-inflated Poisson (ZIP), or zero-inflated
negative binomial (ZINB).


# Data hierarchy

Correct specification of the sampling hierarchy is essential.

A typical eDNA study may contain sites, biological samples nested within sites,
and technical or sequencing replicates nested within biological samples.

For example:

```text
Site
├── Biological sample 1
│   ├── Technical replicate 1
│   ├── Technical replicate 2
│   └── Technical replicate 3
│
├── Biological sample 2
│   ├── Technical replicate 1
│   ├── Technical replicate 2
│   └── Technical replicate 3
│
└── Biological sample 3
    ├── Technical replicate 1
    ├── Technical replicate 2
    └── Technical replicate 3
```

The corresponding `FitModel_joint()` arguments are:

- `site_col`: column identifying the occupancy unit;
- `sample_col`: column identifying the biological sample;
- `replicate_col`: optional column identifying technical or sequencing
  replicates;
- `otu_col`: OTU identifier in the internally prepared long-format data;
- `count_col`: observed sequencing count.

For example:

```{r hierarchy-example, eval=FALSE}
fit <- FitModel_joint(
  phyloseq = ps,
  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate"
)
```

The choice of `site_col` is scientifically important because it defines the
level at which the latent occupancy state is estimated.

For example, if occupancy represents presence at a location during a
particular sampling month, a location-by-month identifier can be used.


# Preparing the example phyloseq data

The package contains example `phyloseq` data that can be used to demonstrate
the model.

```{r prepare-data, message=FALSE, warning=FALSE}
library(phyloseq)
library(eDNAModel)
library(dplyr)

data("physeq_new", package = "eDNAModel")

ps <- physeq_new$`Marine invasive species Thessaloniki`
```

The sample metadata can be inspected using:

```{r inspect-data}
sample_variables(ps)

head(
  as.data.frame(
    phyloseq::sample_data(ps)
  )
)
```

For this example, an occupancy unit is defined by the combination of sampling
area and sampling month.

```{r create-site-month}
meta <- as.data.frame(
  phyloseq::sample_data(ps)
)

meta$site_month <- interaction(
  meta$Sampling.area.Name,
  meta$Samplingmonth,
  drop = TRUE,
  sep = "_"
)

phyloseq::sample_data(ps) <-
  phyloseq::sample_data(meta)
```

The resulting hierarchy can be inspected with:

```{r inspect-hierarchy}
meta |>
  dplyr::summarise(
    n_rows = dplyr::n(),
    n_sites = dplyr::n_distinct(siteCode),
    n_site_month = dplyr::n_distinct(site_month),
    n_samples = dplyr::n_distinct(Name),
    n_replicates = dplyr::n_distinct(Replicate)
  )
```

Before fitting a model, verify that:

- counts are non-negative;
- sample identifiers are correctly defined;
- the site variable represents the intended occupancy unit;
- biological samples are correctly identified;
- technical replicates are correctly identified when present;
- model covariates do not contain unintended missing values.


# Fitting a basic joint model

The simplest model contains intercepts for all three processes.

```{r basic-model, eval=FALSE}
fit <- FitModel_joint(
  phyloseq = ps,

  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",

  occupancy_formula = ~ 1,
  capture_formula = ~ 1,
  abundance_formula = ~ 1,

  abundance_family = "poisson"
)
```

This model estimates overall occupancy, capture, and abundance components while
allowing the default random-effect structure implemented by
`FitModel_joint()`.


# Fixed-effect interpretation

The estimated fixed parameters are stored in:

```{r fixed-effects, eval=FALSE}
fit$fixed_effects
```

For an intercept-only occupancy model, the occupancy intercept can be
transformed to the probability scale.

```{r occupancy-probability, eval=FALSE}
beta_occ <- fit$fixed_effects |>
  dplyr::filter(parameter == "beta_occ") |>
  dplyr::pull(Estimate)

psi <- plogis(beta_occ)

psi
```

Thus,

\[
\psi = \mathrm{logit}^{-1}(\beta_{\mathrm{occ}}).
\]


## Capture probability

The capture intercept can be transformed in the same way:

```{r capture-probability, eval=FALSE}
beta_cap <- fit$fixed_effects |>
  dplyr::filter(parameter == "beta_cap") |>
  dplyr::pull(Estimate)

p_capture <- plogis(beta_cap)

p_capture
```

Thus,

\[
p = \mathrm{logit}^{-1}(\beta_{\mathrm{cap}}).
\]


## Conditional abundance

For a log-linked abundance model:

```{r abundance-mean, eval=FALSE}
beta_abund <- fit$fixed_effects |>
  dplyr::filter(parameter == "beta_abund") |>
  dplyr::pull(Estimate)

lambda <- exp(beta_abund)

lambda
```

Therefore,

\[
\lambda = \exp(\beta_{\mathrm{abund}}).
\]

For an intercept-only model, this gives the conditional expected count
associated with the abundance intercept, subject to the remaining model
structure.


# Adding covariates

Different predictors can be assigned to different stages of the hierarchical
process.

For example:

```{r covariate-model, eval=FALSE}
fit_cov <- FitModel_joint(
  phyloseq = ps,

  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",

  occupancy_formula = ~ Samplingmonth,
  capture_formula = ~ 1,
  abundance_formula = ~ Samplingmonth + Replicate,

  abundance_family = "nbinom"
)
```

This specification allows sampling month to explain variation in occupancy and
conditional abundance, while `Replicate` is included in the abundance
component.

More generally, covariates should be assigned according to the process they
are expected to influence.

For example:

- environmental variables may explain occupancy;
- sampling-effort variables may explain capture;
- sequencing-depth or laboratory variables may explain observed read
  abundance.

The appropriate assignment depends on the scientific design of the study.


# Abundance families

The conditional abundance distribution is controlled using
`abundance_family`.

Four families are available in the current joint modelling framework.


## Poisson

```{r poisson-model, eval=FALSE}
fit_poisson <- FitModel_joint(
  phyloseq = ps,

  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",

  occupancy_formula = ~ 1,
  capture_formula = ~ 1,
  abundance_formula = ~ 1,

  abundance_family = "poisson"
)
```

The Poisson model is the simplest count model and assumes that the conditional
variance is determined by the conditional mean.


## Negative binomial

```{r nb-model, eval=FALSE}
fit_nb <- FitModel_joint(
  phyloseq = ps,

  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",

  occupancy_formula = ~ 1,
  capture_formula = ~ 1,
  abundance_formula = ~ 1,

  abundance_family = "nbinom"
)
```

The negative-binomial family introduces an additional dispersion parameter and
can accommodate conditional overdispersion relative to the Poisson model.


## Zero-inflated Poisson

```{r zip-model, eval=FALSE}
fit_zip <- FitModel_joint(
  phyloseq = ps,

  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",

  occupancy_formula = ~ 1,
  capture_formula = ~ 1,
  abundance_formula = ~ 1,

  abundance_family = "zip"
)
```


## Zero-inflated negative binomial

```{r zinb-model, eval=FALSE}
fit_zinb <- FitModel_joint(
  phyloseq = ps,

  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",

  occupancy_formula = ~ 1,
  capture_formula = ~ 1,
  abundance_formula = ~ 1,

  abundance_family = "zinb"
)
```

ZIP and ZINB introduce an additional zero-inflation component.

These models should not be selected simply because the raw OTU table contains
many zeros. Occupancy and capture already provide mechanisms that generate
zeros.

A zero-inflated conditional count distribution is most useful when additional
zero inflation remains in the conditional abundance process.


# Random effects

`FitModel_joint()` provides controls for several Gaussian random-effect
components.

These include:

- `random_occ_otu`;
- `random_capture_otu`;
- `random_abund_otu`;
- `random_sample`;
- `random_sample_otu`;
- `random_zi_otu`.

For example:

```{r random-effects-model, eval=FALSE}
fit_re <- FitModel_joint(
  phyloseq = ps,

  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",

  occupancy_formula = ~ 1,
  capture_formula = ~ 1,
  abundance_formula = ~ 1,

  random_occ_otu = TRUE,
  random_capture_otu = TRUE,
  random_abund_otu = TRUE,

  random_sample = TRUE,
  random_sample_otu = FALSE,

  abundance_family = "nbinom"
)
```

Random-effect standard deviations are represented internally on the log-SD
scale.

For example, if

\[
\log(\sigma)=-0.69,
\]

then

\[
\sigma=\exp(-0.69)\approx0.50.
\]

The `random_zi_otu` component is relevant to zero-inflated models.


# OTU filtering

Rare OTUs can be filtered before model fitting using
`min_species_sum` and `min_detection_replicates`.

For example:

```{r filtering, eval=FALSE}
fit_filtered <- FitModel_joint(
  phyloseq = ps,

  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",

  occupancy_formula = ~ 1,
  capture_formula = ~ 1,
  abundance_formula = ~ 1,

  abundance_family = "nbinom",

  min_species_sum = 10,
  min_detection_replicates = 2
)
```

`min_species_sum` controls the minimum total count required for an OTU to be
retained.

`min_detection_replicates` controls the minimum number of positive
observations required for an OTU.

The retained taxa and filtering information can be inspected using:

```{r filtering-output, eval=FALSE}
length(fit_filtered$retained_otus)

fit_filtered$retained_otus

fit_filtered$otu_stats
```

Filtering decisions should be reported because they determine which taxa
contribute to model estimation.


# Abundance offsets

An exposure variable can be incorporated into the abundance component using
`abundance_offset`.

Conceptually, an offset modifies the abundance predictor as

\[
\log(\lambda)
=
\mathbf{x}^{T}\boldsymbol{\beta}
+
\log(E),
\]

where \(E\) represents exposure.

For eDNA data, an appropriate exposure could be a positive variable
representing differential sampling or sequencing effort, provided its use is
scientifically justified.

For example:

```{r offset-model, eval=FALSE}
fit_offset <- FitModel_joint(
  phyloseq = ps,

  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",

  occupancy_formula = ~ 1,
  capture_formula = ~ 1,
  abundance_formula = ~ Samplingmonth,

  abundance_offset = "SequencingDepth",

  abundance_family = "nbinom"
)
```

The supplied offset variable represents exposure. In the current
`FitModel_joint()` implementation, the exposure is transformed to the log
scale internally.

An offset should therefore be strictly positive and should represent exposure
rather than an ordinary explanatory covariate.


# Numerical convergence

Numerical convergence should always be evaluated before interpreting model
parameters.

Important convergence information is stored in:

```{r convergence-output, eval=FALSE}
fit$convergence$overall_status

fit$convergence$max_abs_gradient

fit$convergence$pd_hessian

fit$convergence$finite_fixed_parameter_standard_errors

fit$convergence$optimizer_code
```

The current implementation distinguishes between strict and marginal gradient
convergence.

| Status | Gradient criterion |
|:---|:---|
| PASS | `max(abs(gradient)) <= 0.001` |
| MARGINAL | `0.001 < max(abs(gradient)) <= 0.005` |
| FAIL | `max(abs(gradient)) > 0.005` or non-finite |

These thresholds are numerical diagnostics rather than universal statistical
criteria.


## Strict and acceptable convergence

Strict convergence requires the relevant numerical checks to pass, including:

- optimizer convergence;
- a finite outer gradient satisfying the strict gradient tolerance;
- successful `TMB::sdreport()`;
- a positive-definite Hessian;
- finite fixed-parameter standard errors.

The corresponding results can be inspected using:

```{r strict-convergence, eval=FALSE}
fit$convergence$strict_convergence

fit$convergence$acceptable_convergence

fit$convergence$overall_status
```

A model may receive an overall status of `"MARGINAL"` when the optimizer,
`sdreport`, Hessian, and standard-error diagnostics pass but the maximum
absolute gradient lies above the strict threshold and within the specified
marginal range.

A marginal fit therefore deserves additional inspection but is not equivalent
to an outright optimizer or Hessian failure.


## Automatic optimization restarts

`FitModel_joint()` can restart optimization from the current solution when
the strict gradient criterion has not been achieved.

For example:

```{r restart-example, eval=FALSE}
fit <- FitModel_joint(
  phyloseq = ps,

  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",

  occupancy_formula = ~ 1,
  capture_formula = ~ 1,
  abundance_formula = ~ 1,

  abundance_family = "nbinom",

  gradient_tol = 1e-3,
  gradient_marginal_factor = 5,

  max_restarts = 2
)
```

With `max_restarts = 2`, the fitting procedure can use up to three optimization
passes.

Optimization history can be inspected using:

```{r optimization-history, eval=FALSE}
fit$convergence$n_optimization_passes

fit$convergence$selected_pass

fit$convergence$optimization_history

fit$convergence$largest_gradients
```

The largest-gradient table can help identify which fixed parameters are
responsible for marginal or failed gradient diagnostics.


# Additional parameter diagnostics

Additional diagnostics are stored in:

```{r parameter-diagnostics, eval=FALSE}
fit$diagnostics$table

fit$diagnostics$near_zero_sd

fit$diagnostics$large_se_parameters
```

A near-zero random-effect variance may indicate that the corresponding source
of heterogeneity is weakly supported by the data.

A large standard error relative to a parameter estimate may indicate
substantial uncertainty or weak identification.

These diagnostics should be considered warnings for further investigation
rather than automatic evidence that the entire model is invalid.


# Memory considerations for large datasets

Joint community models can become computationally demanding because the
likelihood may contain observations across many sites, biological samples,
technical replicates, and OTUs.

For large datasets, expensive covariance calculations can be disabled.

```{r memory-model, eval=FALSE}
fit_large <- FitModel_joint(
  phyloseq = ps,

  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",

  occupancy_formula = ~ 1,
  capture_formula = ~ 1,
  abundance_formula = ~ 1,

  abundance_family = "nbinom",

  get_report_covariance = FALSE,
  get_joint_precision = FALSE
)
```

Setting `get_report_covariance = FALSE` can substantially reduce memory use
when a large covariance matrix of reported quantities is not needed.

Setting `get_joint_precision = FALSE` can further reduce memory requirements
when the joint precision matrix is not required.


# Examining fitted-model output

A fitted model returned by `FitModel_joint()` is an object of class
`eDNAModel_joint`.

```{r fitted-class, eval=FALSE}
class(fit)
```

Important components can include:

```{r fitted-components, eval=FALSE}
fit$fixed_effects

fit$derived

fit$site_data

fit$sample_data

fit$long_df

fit$otu_stats

fit$retained_otus

fit$formulas

fit$abundance_family

fit$convergence

fit$diagnostics

fit$random_effects
```

Approximate 95% Wald confidence intervals for fixed parameters can be
constructed as:

```{r confidence-intervals, eval=FALSE}
results <- fit$fixed_effects |>
  dplyr::mutate(
    lower = Estimate - 1.96 * `Std. Error`,
    upper = Estimate + 1.96 * `Std. Error`
  )

results
```

These intervals should be interpreted only after checking the numerical
diagnostics and considering whether the corresponding parameters are
adequately identified.


# Interpreting zeros in eDNA data

A key motivation for the joint framework is that a zero count does not
necessarily imply ecological absence.

A zero may occur because:

1. the OTU is genuinely absent from the occupancy unit;
2. the OTU is present but is not captured in a particular biological sample;
3. the OTU passes the occupancy and capture stages but the conditional count
   process produces a zero;
4. for ZIP or ZINB models, an additional zero arises from the zero-inflation
   component.

The hierarchical model attempts to separate these mechanisms statistically.

This distinction is particularly important in metabarcoding studies, where
imperfect capture and heterogeneous sequencing counts can otherwise be
confounded with ecological occurrence.


# Choosing among Poisson, NB, ZIP, and ZINB

The four abundance families represent different assumptions about the
conditional count process.

A useful strategy is to begin with a parsimonious model and introduce
additional complexity only when scientifically and statistically justified.

For example:

```{r compare-families, eval=FALSE}
fit_poisson <- FitModel_joint(
  phyloseq = ps,
  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",
  abundance_family = "poisson"
)

fit_nb <- FitModel_joint(
  phyloseq = ps,
  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",
  abundance_family = "nbinom"
)

fit_zip <- FitModel_joint(
  phyloseq = ps,
  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",
  abundance_family = "zip"
)

fit_zinb <- FitModel_joint(
  phyloseq = ps,
  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",
  abundance_family = "zinb"
)
```

The overall proportion of zeros should not, by itself, determine whether a
zero-inflated family is required.

Because occupancy and capture already generate zeros, ZIP or ZINB models
represent additional zero inflation in the conditional count component.


# Recommended modelling workflow

A practical workflow is:

1. Verify the sampling hierarchy.
2. Inspect the count distribution and sparsity.
3. Define the occupancy unit.
4. Define biological samples and technical replicates.
5. Select biologically meaningful occupancy covariates.
6. Select sampling-related capture covariates.
7. Select abundance covariates and any scientifically justified exposure
   offset.
8. Choose an initial abundance family.
9. Fit the model.
10. Inspect optimizer convergence.
11. Inspect the maximum absolute gradient.
12. Check the Hessian and standard errors.
13. Inspect random-effect variance estimates.
14. Compare scientifically motivated alternative specifications.
15. Evaluate model adequacy using simulation or external validation where
    appropriate.
16. Report both parameter estimates and numerical diagnostics.


# Complete example

The following example illustrates a ZINB model using the Thessaloniki dataset.

The chunk is not evaluated while building the vignette because fitting a
joint TMB model can be computationally expensive.

```{r complete-example, eval=FALSE}
library(phyloseq)
library(eDNAModel)
library(dplyr)

# Load example data
data("physeq_new", package = "eDNAModel")

ps <- physeq_new$`Marine invasive species Thessaloniki`

# Extract sample metadata
meta <- as.data.frame(
  phyloseq::sample_data(ps)
)

# Define occupancy unit:
# sampling location x sampling month
meta$site_month <- interaction(
  meta$Sampling.area.Name,
  meta$Samplingmonth,
  drop = TRUE,
  sep = "_"
)

# Return metadata to phyloseq object
phyloseq::sample_data(ps) <-
  phyloseq::sample_data(meta)

# Fit ZINB joint model
fit_zinb <- FitModel_joint(
  phyloseq = ps,

  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",

  occupancy_formula = ~ 1,
  capture_formula = ~ 1,
  abundance_formula = ~ 1,

  abundance_family = "zinb",

  gradient_tol = 1e-3,
  gradient_marginal_factor = 5,

  max_restarts = 2,

  get_report_covariance = FALSE,
  get_joint_precision = FALSE,

  verbose = TRUE,

  tmb_verbose = FALSE
)
```

After fitting, parameter estimates can be inspected using:

```{r complete-output, eval=FALSE}
fit_zinb$fixed_effects
```

Overall numerical convergence status:

```{r complete-convergence, eval=FALSE}
fit_zinb$convergence$overall_status
```

Maximum absolute gradient:

```{r complete-gradient, eval=FALSE}
fit_zinb$convergence$max_abs_gradient
```

Positive-definite Hessian:

```{r complete-hessian, eval=FALSE}
fit_zinb$convergence$pd_hessian
```

Full diagnostic table:

```{r complete-diagnostics, eval=FALSE}
fit_zinb$diagnostics$table
```


# Reproducibility and reporting

For reproducible analyses, report at least:

- the definition of the occupancy unit;
- the biological-sample and technical-replicate hierarchy;
- the occupancy formula;
- the capture formula;
- the abundance formula;
- the abundance family;
- OTU-filtering thresholds;
- enabled random effects;
- the abundance offset, if one is used;
- optimizer and gradient settings when changed from defaults;
- the final convergence status;
- the maximum absolute gradient;
- whether the Hessian was positive definite;
- whether fixed-parameter standard errors were finite.

The fitted object stores model specifications and numerical diagnostics that
can be used to document the analysis after fitting.


# Summary

`FitModel_joint()` provides a hierarchical framework for community-level eDNA
count data in which ecological occurrence, biological capture, and sequencing
abundance are represented as distinct components of a joint likelihood.

The framework is particularly useful when:

- species detection is imperfect;
- several biological samples are collected within occupancy units;
- technical or sequencing replicates are available;
- read abundance is overdispersed;
- OTUs differ substantially in occupancy, capture, or abundance;
- additional conditional zero inflation may be present;
- community-level inference requires sharing information across OTUs.

Careful specification of the sampling hierarchy is essential because
`site_col`, `sample_col`, and `replicate_col` correspond to different levels of
the ecological and observation process.

Likewise, numerical convergence should be evaluated systematically before
interpreting parameter estimates. Optimizer status, gradients, Hessian
diagnostics, standard errors, random-effect variances, and parameter
uncertainty provide complementary information about model reliability.
