---
title: "eDNAModel: Small Example Workflow"
author: "Marwah Soliman, Bert van der Veen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{eDNAModel: Small Example Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup1, include=FALSE}
library(eDNAModel)
library(phyloseq)
```


## Introduction

The `eDNAModel` package provides a pipeline to analyze environmental DNA (eDNA) data for species occupancy and abundance modeling using TMB.

This vignette demonstrates:

1. Loading a phyloseq object
2. Converting to a 3D array
3. Filtering the array
4. Running the TMB pipeline

---

## Load a phyloseq object

We'll use an example `.RDS` file located in `inst/extdata/`.

```{r load-data, eval = TRUE}
# Load example phyloseq object
data_path <- system.file("extdata", "longdataexample.RDS", package = "eDNAModel")
data_array <- data_array_phyloseq(data_path, verbose = FALSE)

# Subset to 10 species, 6 sites, and 3 replicates
small_array <- data_array[1:10, 1:6, 1:3]

# Filter species with at least 10 total counts
filtered_array <- filter_data_array(small_array, min_species_sum = 10, save_path = NULL)

# Show dimensions
dim(filtered_array)
```

---

## Filter the 3D Data Array

We remove species with all-zero counts, sites with all-zero counts, and rare species.

```{r input, eval = TRUE}
Y_small <- to2D(filtered_array)
X_small <- Y_small[, c("Site", "Replicate")]
X_small$Site <- factor(X_small$Site)
X_small$Replicate <- factor(X_small$Replicate)
```

---

## Run the Full TMB Pipeline

Fit the statistical model to the filtered array:

```{r model, eval = TRUE}
model_small <- run_full_TMB(
  data_array_filtered = filtered_array,
  X = X_small,
  a.formula = ~ Site + (1 | Replicate),
  o.formula = ~ Site,
  linko = 1,  # logit
  linka = 0,  # log
  family = 1  # ZINB
)
```

```{r sum1, echo=FALSE}
summary_output <- summary_all(model_small)
head(summary_output)
```

```{r fitted_summary, echo=FALSE}
fitted_vals <- fitted_TMB(model_small)

# Summary of fitted values
summary(fitted_vals$fitted_occupancy)
summary(fitted_vals$fitted_abundance)
```

```{r run_resid, eval = TRUE}
# Residuals
residuals_out <- compute_residuals_TMB(model_small, filtered_array,type="pearson")
hist(residuals_out$abundance_residuals, main = "Abundance Residuals")
hist(residuals_out$occupancy_residuals, main = "Occupancy Residuals")
```


```{r run-pred, eval = TRUE}
newX <- unique(X_small[, "Site", drop = FALSE])
newX$Site <- factor(newX$Site)

pred_abund <- predict_TMB(model_small, newX = newX, formula = ~ Site, which = "abundance",type = ("response"))
pred_occ <- predict_TMB(model_small, newX = newX, formula = ~ Site, which = "occupancy",type = ("response"))

head(pred_abund)
head(pred_occ)
```

```{r run-H, eval = TRUE}
# Check for extreme occupancy values
high_occ <- fitted_vals$fitted_occupancy >= 0.999
low_occ  <- fitted_vals$fitted_occupancy <= 0.001

if (any(high_occ)) {
  warning(sum(high_occ), " entries with occupancy ~1. Possible overfitting.")
}
if (any(low_occ)) {
  warning(sum(low_occ), " entries with occupancy ~0. Check sparsity.")
}

```

You can extract model output such as slopes, intercepts, and model fit metrics from `result`.

---

## Summary

This vignette showed how to:

- Convert phyloseq data to 3D array
- Filter the array
- Run TMB model pipeline

Explore the `eDNAModel` documentation for further customization and model diagnostics.
