---
title: "eDNAModel: Updated Workflow with Phyloseq"
author: "Marwah Soliman, Bert van der Veen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{eDNAModel: Updated Workflow with Phyloseq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup1, include=FALSE}
library(eDNAModel)
library(phyloseq)
```


## Introduction

The eDNAModel package provides a robust pipeline for analyzing environmental DNA (eDNA) using hierarchical multispecies occupancy-abundance models implemented via Template Model Builder (TMB).

1.This updated vignette demonstrates how to:

2.Load a phyloseq object

3.Automatically prepare the data using fit.phyloseq()

4.Fit the model with minimal effort

5.Extract and visualize residuals and predictions

---

## Load a phyloseq object

We'll use an example `.RDS` file located in `inst/extdata/`.

```{r load-data, eval = TRUE}
physeq_path <- system.file("extdata", "longdataexample.RDS", package = "eDNAModel")
physeq <- readRDS(physeq_path)
```

---

## Run the Full TMB Pipeline
We now use the high-level wrapper fit.phyloseq() which handles:

Extracting OTU table

Filtering biological samples

Adding Site, Sample, Replicate columns

Running the TMB pipeline


```{r model, eval = TRUE}
model <- fit.phyloseq(
  phyloseq_obj = physeq,
  a.formula = ~ Site + diag(1 | Sample) + diag(1 | Replicate),
  o.formula = ~ Site,
  linko = 1,
  linka = 0,
  family = 1,
  control = list(startOptcontrol = list(maxit = 200),
  optControl = list(maxit = 10e3, sigma1 = 0.25), trace = TRUE)
)
```
---

##Model Summary

```{r sum1, echo=FALSE}
summary_out <- summary(model)
head(summary_out)
```

```{r exp, echo=FALSE}
attr(summary_out, "explanation")
```
---

## Fitted Values

```{r fitted_summary, echo=FALSE}
fitted_vals <- fitted_TMB(model)

summary(fitted_vals$fitted_abundance)
summary(fitted_vals$fitted_occupancy)

```
---

##Residuals

```{r run_resid, eval = TRUE}
# Residuals
residuals <- compute_residuals_TMB(model = model, y = y, X = X, type = "pearson")

hist(residuals$abundance_residuals, main = "Abundance Residuals", col = "lightblue")
hist(residuals$occupancy_residuals, main = "Occupancy Residuals", col = "lightgreen")

```
---

## Predictions

```{r run-pred, eval = TRUE}
newdata <- unique(X[, "Site", drop = FALSE])
newdata$Site <- factor(newdata$Site)

pred_abund <- predict_TMB(model, newX = newdata, formula = ~ Site, which = "abundance", type = "response")
pred_occ <- predict_TMB(model, newX = newdata, formula = ~ Site, which = "occupancy", type = "response")

head(pred_abund)
head(pred_occ)


```
---

##Check for Boundary Cases

```{r run-H, eval = TRUE}
# Check for extreme occupancy values
high_occ <- fitted_vals$fitted_occupancy >= 0.999
low_occ <- fitted_vals$fitted_occupancy <= 0.001

if (any(high_occ)) {
  warning(sum(high_occ), " entries with occupancy ~1. Possible overfitting.")
}
if (any(low_occ)) {
  warning(sum(low_occ), " entries with occupancy ~0. Check sparsity.")
}

```
---

##  Hessian Diagnostic

```{r run-Hes, eval = TRUE}
H <- tryCatch({
  model$TMBobj$he()
}, error = function(e) {
  message("ℹ️ Hessian calculation skipped (random effects present).")
  NULL
})

if (!is.null(H)) {
  eigenvalues <- eigen(H)$values
  if (any(eigenvalues <= 0)) {
    warning("⚠️ Non-positive definite Hessian detected. Consider simplifying the model.")
  } else {
    message("✅ Hessian is positive definite.")
  }
}

```
---

You can extract model output such as slopes, intercepts, and model fit metrics from `result`.

---

## Summary

This vignette demonstrated how to:

-- Use fit.phyloseq() for end-to-end modeling

-- Interpret and visualize model output

-- Compute residuals and predictions

--Perform diagnostics

Explore the `eDNAModel` documentation for further customization and model diagnostics.
