---
title: "eDNAModel: gllvm"
author: "Marwah Soliman, Bert van der Veen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{eDNAModel: gllvm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load-packages, message=FALSE, warning=FALSE}
library(phyloseq)
library(Matrix)
library(eDNAModel)
library(dplyr)
library(tidyr)
library(ggplot2)
library(pheatmap)
library(tibble)
library(gllvm)
library(glmmTMB)
library(reshape2)
library(grid)

```

# 1. Introduction

This vignette demonstrates the updated occupancy–detection model workflow using the `eDNAModel` package and `phyloseq` data objects. The workflow includes fitting binomial and Poisson mixed-effects models across iterations, visualizing occupancy and detection probabilities, using gllvm for occupancy model.

---

# 2. Load and Subset Data
```{r load-data}
physeq_new=readRDS("~/Downloads/BGE_Port_sampling.rds")

data("physeq_new",package = "eDNAModel")

physeq_one=physeq_new$`Marine Invasive species Trondheim`
```

# 3. Fit the Occupancy–Detection Model

```{r fit-model, echo=FALSE}
out <- FitModel_gllvm(
  phyloseq = physeq_one,
  site_col = "Name",

  abundance_rhs =
    (1 | Samplingmonth / OTU) +
    (1 | OTU),

  occupancy_covars = "Samplingmonth",

  abundance_family = "poisson",

  n_iter = 50,
  burn_in = 10,
  num_lv_c = 2
)

```
# 4. Visualization

## 4.1 Occupancy Probability Caterpillar Plot

```{r plot-occupancy, fig.width=10, fig.height=9}

library(dplyr)
library(ggplot2)

otu_col <- "OTU"

# ------------------------------------------------------------
# 1. Combine retained GLLVM occupancy predictions
# ------------------------------------------------------------

psi_draws <- dplyr::bind_rows(out$psi_list)

# eta is on the logit scale
psi_draws <- psi_draws %>%
    dplyr::mutate(
        psi = stats::plogis(.data$eta)
    )

# ------------------------------------------------------------
# 2. Summarise occupancy at the OTU level
#
# This averages across sites and retained iterations.
# ------------------------------------------------------------

psi_otu_summary <- psi_draws %>%
    dplyr::group_by(.data[[otu_col]]) %>%
    dplyr::summarise(
        psi_mean = mean(.data$psi, na.rm = TRUE),
        psi_median = stats::median(.data$psi, na.rm = TRUE),
        psi_lwr = stats::quantile(
            .data$psi,
            0.025,
            na.rm = TRUE,
            names = FALSE
        ),
        psi_upr = stats::quantile(
            .data$psi,
            0.975,
            na.rm = TRUE,
            names = FALSE
        ),
        .groups = "drop"
    ) %>%
    dplyr::arrange(.data$psi_mean)

# ------------------------------------------------------------
# 3. Order OTUs by occupancy
# ------------------------------------------------------------

psi_otu_summary[[otu_col]] <- factor(
    psi_otu_summary[[otu_col]],
    levels = psi_otu_summary[[otu_col]]
)

# ------------------------------------------------------------
# 4. Caterpillar plot
# ------------------------------------------------------------

ggplot(
    psi_otu_summary,
    aes(
        x = psi_mean,
        y = .data[[otu_col]]
    )
) +
    geom_point(
        color = "darkgreen",
        size = 2
    ) +
    geom_errorbarh(
        aes(
            xmin = psi_lwr,
            xmax = psi_upr
        ),
        height = 0.2,
        color = "darkgreen"
    ) +
    labs(
        title = "Caterpillar Plot: Occupancy Probability by OTU",
        x = expression("Occupancy probability (" * psi * ")"),
        y = "OTU"
    ) +
    theme_minimal() +
    theme(
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5)
    )


```
## 4.2 Detection Probability Caterpillar Plot

```{r plot-detection,fig.width=10, fig.height=9}
library(dplyr)
library(ggplot2)

otu_col <- "OTU"

# ------------------------------------------------------------
# 1. Combine retained detection predictors
# ------------------------------------------------------------

p_detect_draws <- dplyr::bind_rows(out$p_detect_list)

# eta is on the cloglog scale
p_detect_draws <- p_detect_draws %>%
    dplyr::mutate(
        p_detect = 1 - exp(-exp(.data$eta))
    )

# ------------------------------------------------------------
# 2. Summarise detection probability by OTU
# ------------------------------------------------------------

p_detect_otu_summary <- p_detect_draws %>%
    dplyr::group_by(.data[[otu_col]]) %>%
    dplyr::summarise(
        p_detect_mean = mean(.data$p_detect, na.rm = TRUE),
        p_detect_median = stats::median(.data$p_detect, na.rm = TRUE),
        p_detect_lwr = stats::quantile(
            .data$p_detect,
            0.025,
            na.rm = TRUE,
            names = FALSE
        ),
        p_detect_upr = stats::quantile(
            .data$p_detect,
            0.975,
            na.rm = TRUE,
            names = FALSE
        ),
        .groups = "drop"
    ) %>%
    dplyr::arrange(.data$p_detect_mean)

# ------------------------------------------------------------
# 3. Order OTUs
# ------------------------------------------------------------

p_detect_otu_summary[[otu_col]] <- factor(
    p_detect_otu_summary[[otu_col]],
    levels = p_detect_otu_summary[[otu_col]]
)

# ------------------------------------------------------------
# 4. Caterpillar plot
# ------------------------------------------------------------

ggplot(
    p_detect_otu_summary,
    aes(
        x = p_detect_mean,
        y = .data[[otu_col]]
    )
) +
    geom_point(
        color = "steelblue",
        size = 2
    ) +
    geom_errorbarh(
        aes(
            xmin = p_detect_lwr,
            xmax = p_detect_upr
        ),
        height = 0.2,
        color = "steelblue"
    ) +
    labs(
        title = "Caterpillar Plot: Detection Probability by OTU",
        x = "Detection Probability",
        y = "OTU"
    ) +
    theme_minimal() +
    theme(
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5)
    )
 
```

## 4.2  biplot

```{r biplot,fig.width=10, fig.height=9}

library(ggplot2)
library(grid)

site_col <- "Name"
otu_col  <- "OTU"

# Extract latent coordinates
site_mat <- as.matrix(
  out$mean_lv_sites[, c("LV1", "LV2")]
)

species_mat <- as.matrix(
  out$mean_lv_species[, c("LV1", "LV2")]
)

# Common symmetric scaling
site_radius <- max(
  sqrt(rowSums(site_mat^2)),
  na.rm = TRUE
)

species_radius <- max(
  sqrt(rowSums(species_mat^2)),
  na.rm = TRUE
)

scale_factor <- sqrt(
  site_radius / species_radius
)

scaled_sites <- site_mat / scale_factor
scaled_species <- species_mat * scale_factor

# Add labels
scaled_sites_df <- data.frame(
  Site = out$mean_lv_sites[[site_col]],
  LV1  = scaled_sites[, 1],
  LV2  = scaled_sites[, 2]
)

scaled_species_df <- data.frame(
  OTU = out$mean_lv_species[[otu_col]],
  LV1 = scaled_species[, 1],
  LV2 = scaled_species[, 2]
)

# Plot
ggplot() +

  geom_point(
    data = scaled_sites_df,
    aes(x = LV1, y = LV2),
    color = "blue",
    size = 2
  ) +

  geom_text(
    data = scaled_sites_df,
    aes(x = LV1, y = LV2, label = Site),
    color = "blue",
    size = 2.5,
    vjust = -0.5
  ) +

  geom_segment(
    data = scaled_species_df,
    aes(
      x = 0,
      y = 0,
      xend = LV1,
      yend = LV2
    ),
    arrow = grid::arrow(
      length = grid::unit(0.2, "cm")
    ),
    color = "brown",
    alpha = 0.8
  ) +

  geom_text(
    data = scaled_species_df,
    aes(
      x = LV1,
      y = LV2,
      label = OTU
    ),
    color = "brown",
    size = 3,
    vjust = -0.5
  ) +

  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = 0.3
  ) +

  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 0.3
  ) +

  coord_equal() +

  labs(
    x = "Latent Variable 1",
    y = "Latent Variable 2",
    title = "GLLVM Latent Variable Biplot"
  ) +

  theme_minimal()

```
