---
title: "eDNAModel: Updated Workflow with Phyloseq"
author: "Marwah Soliman, Bert van der Veen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{eDNAModel: Updated Workflow with Phyloseq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load-packages, message=FALSE, warning=FALSE}
library(phyloseq)
library(Matrix)
library(eDNAModel)
library(dplyr)
library(tidyr)
library(ggplot2)
library(pheatmap)
library(tibble)
library(gllvm)
library(glmmTMB)
library(reshape2)
```

# 1. Introduction

This vignette demonstrates the updated occupancy–detection model workflow using the `eDNAModel` package and `phyloseq` data objects. The workflow includes fitting binomial and Poisson mixed-effects models across iterations, visualizing occupancy and detection probabilities, and comparing models using AIC/BIC.

---

# 2. Load and Subset Data
```{r load-data}
#physeq_new=readRDS("~/Downloads/BGE_Port_sampling.rds")

data("physeq_new",package = "eDNAModel")

physeq_one=physeq_new$`Marine invasive species Thessaloniki`

#physeq_path <- system.file("extdata", "longdataexample.RDS", package = "eDNAModel")
#physeq <- readRDS(physeq_path)

# Subset 100 random OTUs
#set.seed(123)
#selected_otus <- sample(taxa_names(physeq), 100)
#physeq_100 <- prune_taxa(selected_otus, physeq)
```

---

# 3. Fit the Occupancy–Detection Model

```{r fit-model, echo=FALSE}
out <- FitModel(
  phyloseq      = ps,
  site_col      = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",
  otu_col = "OTU",
  count_col = "y",
  abundance_formula =  y ~  (1 | OTU) + (1 | Name / OTU),
  occupancy_formula  = z_sim ~(1 | OTU),
  capture_formula =    a_sim ~ 1 + (1 | OTU),
  min_species_sum = 10,
  n_iter = 20, burn_in = 5,
  abundance_family = "nbinom"
)


```

---

# 4. Visualization

## 4.1 Occupancy Probability Caterpillar Plot

```{r plot-occupancy, fig.width=10, fig.height=9}

library(dplyr)
library(ggplot2)

# Combine posterior samples (after burn-in)
psi_draws <- bind_rows(out$psi_list)

# Summarize properly (quantile-based)
psi_otu_summary <- psi_draws %>%
    group_by(OTU) %>%
    summarise(
        psi_mean = mean(plogis(eta), na.rm = TRUE),
        psi_lwr  = quantile(plogis(eta), 0.025, na.rm = TRUE),
        psi_upr  = quantile(plogis(eta), 0.975, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    arrange(psi_mean) %>%
    mutate(OTU = factor(OTU, levels = OTU))

# Plot
ggplot(psi_otu_summary, aes(x = psi_mean, y = OTU)) +
    geom_point(color = "darkgreen", size = 2) +
    geom_errorbarh(
        aes(xmin = psi_lwr, xmax = psi_upr),
        height = 0.2,
        color = "darkgreen"
    ) +
    labs(
        title = "Caterpillar Plot: Occupancy Probability by OTU",
        x = "Occupancy Probability (ψ)",
        y = "OTU"
    ) +
    theme_minimal() +
    theme(
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5)
    )


```

---

## 4.2 Detection Probability Caterpillar Plot

```{r plot-detection, fig.width=10, fig.height=9}

library(dplyr)
library(ggplot2)

# Combine iteration draws
p_detect_draws <- bind_rows(out$p_detect_list)

# Proper summary (quantile-based)
p_detect_otu_summary <- p_detect_draws %>%
    group_by(OTU) %>%
    summarise(
        p_detect_mean = mean(1 - exp(-exp(eta)), na.rm = TRUE),
        p_detect_lwr  = quantile(1 - exp(-exp(eta)), 0.025, na.rm = TRUE),
        p_detect_upr  = quantile(1 - exp(-exp(eta)), 0.975, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    arrange(p_detect_mean) %>%
    mutate(OTU = factor(OTU, levels = OTU))

# Plot
ggplot(p_detect_otu_summary, aes(x = p_detect_mean, y = OTU)) +
    geom_point(color = "steelblue", size = 2) +
    geom_errorbarh(
        aes(xmin = p_detect_lwr, xmax = p_detect_upr),
        height = 0.2,
        color = "steelblue"
    ) +
    labs(
        title = "Caterpillar Plot: Detection Probability by OTU",
        x = "Detection Probability (p)",
        y = "OTU"
    ) +
    theme_minimal() +
    theme(
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5)
    )

```

---

## 4.3 Occupancy by treatment

```{r plot-psi-treatment, fig.width=10, fig.height=9}

library(dplyr)
library(ggplot2)

treatment_col <- "Samplingmonth"

# Combine draws
psi_draws <- bind_rows(out$psi_list)

# Merge treatment info from long_df
psi_draws <- psi_draws %>%
    left_join(
        out$long_df %>%
            select(OTU, !!sym(treatment_col)) %>%
            distinct(),
        by = "OTU"
    )

# Proper summary
psi_data <- psi_draws %>%
    group_by(OTU, .data[[treatment_col]]) %>%
    summarise(
        psi_mean = mean(plogis(eta), na.rm = TRUE),
        psi_lwr  = quantile(plogis(eta), 0.025, na.rm = TRUE),
        psi_upr  = quantile(plogis(eta), 0.975, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    group_by(OTU) %>%
    mutate(otu_mean = mean(psi_mean, na.rm = TRUE)) %>%
    ungroup() %>%
    arrange(otu_mean) %>%
    mutate(OTU = factor(OTU, levels = unique(OTU)))

# Plot
ggplot(psi_data, aes(x = psi_mean, y = OTU, color = .data[[treatment_col]])) +
    geom_point(position = position_dodge(width = 0.5)) +
    geom_errorbarh(
        aes(xmin = psi_lwr, xmax = psi_upr),
        position = position_dodge(width = 0.5),
        height = 0.2
    ) +
    labs(
        title = "Caterpillar Plot: Occupancy by OTU and Treatment",
        x = "Occupancy Probability (ψ)",
        y = "OTU",
        color = treatment_col
    ) +
    theme_minimal() +
    theme(
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5)
    )

```

---

## 4.4 Clustered Heatmap of Occupancy Probabilities

```{r heatmap-psi, fig.width=10, fig.height=9}

library(dplyr)
library(tidyr)
library(tibble)
library(pheatmap)

site_col <- "Sampling.area.Name"

# ------------------------------------------------------------
# Step 1: Combine ψ draws
# ------------------------------------------------------------
psi_draws <- bind_rows(out$psi_list)

# ------------------------------------------------------------
# Step 2: Create site–OTU structure
# ------------------------------------------------------------
site_otu <- out$long_df %>%
  select(OTU, !!sym(site_col)) %>%
  distinct()

# ------------------------------------------------------------
# Step 3: Safe join (explicit many-to-many)
# ------------------------------------------------------------
psi_long <- site_otu %>%
  left_join(
    psi_draws,
    by = "OTU",
    relationship = "many-to-many"   
  )

# ------------------------------------------------------------
# Step 4: Compute ψ and aggregate
# ------------------------------------------------------------
psi_matrix <- psi_long %>%
  mutate(psi = plogis(eta)) %>%
  group_by(.data[[site_col]], OTU) %>%
  summarise(
    psi = mean(psi, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = OTU,
    values_from = psi
  ) %>%
  column_to_rownames(site_col) %>%
  as.matrix()

# ------------------------------------------------------------
# Step 5: Handle missing values
# ------------------------------------------------------------
psi_matrix[is.na(psi_matrix)] <- 0

# ------------------------------------------------------------
# Step 6: Plot heatmap
# ------------------------------------------------------------
pheatmap(
  psi_matrix,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  scale = "none",
  color = colorRampPalette(c("white", "darkblue"))(100),
  main = "Heatmap of Occupancy Probability (ψ): Sites × OTUs",
  fontsize_row = 8,
  fontsize_col = 7,
  angle_col = 45
)

```


```{r model-comparison-full, echo=FALSE}

##For example 

results <- compare_models_cv(
  phyloseq = ps,
  site_col = "site_month",
  K = 5,

  models = list(

    A = function(ps) FitModel(
      phyloseq = ps,
      site_col = "site_month",
      sample_col = "Name",
      replicate_col = "Replicate",
      otu_col = "OTU",
      count_col = "y",
      occupancy_formula = z_sim ~ 1 + (1 | OTU),
      capture_formula   = a_sim ~ 1 + (1 | OTU),
      abundance_formula =
        y ~ offset(log(total_reads)) + (1 | OTU) + (1 | Name / OTU),
      n_iter = 12,
      burn_in = 4,
      abundance_family = "nbinom"
    ),

    B = function(ps) FitModel(
      phyloseq = ps,
      site_col = "site_month",
      sample_col = "Name",
      replicate_col = "Replicate",
      otu_col = "OTU",
      count_col = "y",
      occupancy_formula = z_sim ~ 1 + (1 | OTU),
      capture_formula   = a_sim ~ 1 + (1 | OTU),
      abundance_formula =
        y ~ offset(log(total_reads)) +
        (1 | OTU) +
        (1 | `Sampling.area.Name` / OTU) +
        (1 | Name / OTU),
      n_iter = 12,
      burn_in = 4,
      abundance_family = "nbinom"
    ),

    C = function(ps) FitModel(
      phyloseq = ps,
      site_col = "site_month",
      sample_col = "Name",
      replicate_col = "Replicate",
      otu_col = "OTU",
      count_col = "y",
      occupancy_formula = z_sim ~ 1 + (1 | OTU),
      capture_formula   = a_sim ~ 1 + (1 | OTU),
      abundance_formula =
        y ~ (1 | OTU) +
        (1 | siteCode / OTU) +
        (1 | Samplingmonth / OTU) +
        (1 | Name / OTU),
      n_iter = 12,
      burn_in = 4,
      abundance_family = "nbinom"
    )
  )
)

print(results)



