---
title: "eDNAModel: Updated Workflow with Phyloseq"
author: "Marwah Soliman, Bert van der Veen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{eDNAModel: Updated Workflow with Phyloseq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load-packages, message=FALSE, warning=FALSE}
library(phyloseq)
library(Matrix)
library(eDNAModel)
library(dplyr)
library(tidyr)
library(ggplot2)
library(pheatmap)
library(tibble)
library(gllvm)
library(glmmTMB)
library(reshape2)
```

# 1. Introduction

This vignette demonstrates the updated occupancy–detection model workflow using the `eDNAModel` package and `phyloseq` data objects. The workflow includes fitting binomial and Poisson mixed-effects models across iterations, visualizing occupancy and detection probabilities and model comparison.
---

# 2. Load and Subset Data
```{r load-data}
#physeq_new=readRDS("~/Downloads/BGE_Port_sampling.rds")

data("physeq_new",package = "eDNAModel")

physeq_one=physeq_new$`Marine invasive species Thessaloniki`

#physeq_path <- system.file("extdata", "longdataexample.RDS", package = "eDNAModel")
#physeq <- readRDS(physeq_path)

# Subset 100 random OTUs
#set.seed(123)
#selected_otus <- sample(taxa_names(physeq), 100)
#physeq_100 <- prune_taxa(selected_otus, physeq)
```

---

#meta <- as.data.frame(
#  phyloseq::sample_data(physeq_one)
#)

# Extract sampling site from Subcode
#meta$site <- trimws(
 # sub("\\s*\\([^)]*\\)\\s*$", "", meta$Subcode)
#)

# Site × month occupancy unit
#meta$site_month <- interaction(
#  meta$site,
 # meta$Samplingmonth,
 # sep = "_",
 # drop = TRUE
#)

# Put metadata back into phyloseq
#phyloseq::sample_data(physeq_one) <- phyloseq::sample_data(meta)

#meta[, c(
#  "Name",
#  "Subcode",
#  "site",
#  "Samplingmonth",
#  "site_month",
#  "Replicate"
#)]

########################## PS data####################

ps <- readRDS("~/Downloads/ps_porto_12S_modelinput.rds")

############################################################
# 3. Fit the Occupancy–Detection Model

```{r fit-model, echo=FALSE}
out <- FitModel(
    phyloseq      = ps,          # your phyloseq object
    site_col      = "site_month",        # site-level variable
    sample_col    = "Name",              # biological sample ID
    replicate_col = "Replicate",         # PCR replicate (optional but recommended)
    
    # --- MODEL STRUCTURE ---
    occupancy_formula =
  z_sim ~ Samplingmonth +
  (1 | site_month) +
  (1 | OTU) +
  (1 | OTU:Samplingmonth),
    capture_formula   = a_sim ~ 1 + (1 | OTU),
    abundance_formula = y ~ (1 | OTU) + (1 | Name),
    
    # --- REQUIRED ---
    otu_col   = "OTU",
    count_col = "y",
    
    # --- FILTERING ---
    min_species_sum = 10,
    min_detection_replicates = 1,
    abundance_threshold = 0,
    
    # --- EM SETTINGS ---
    n_iter  = 10,
    burn_in = 1,
    
    # --- DISTRIBUTION ---
    abundance_family = "nbinom",
    
    verbose = TRUE, 
    n_sim_ci = 50
)


```

---

# 4. Visualization

## 4.1 Occupancy Probability Caterpillar Plot

```{r plot-occupancy, fig.width=10, fig.height=9}

library(dplyr)
library(ggplot2)

# Combine posterior samples (after burn-in)
psi_draws <- bind_rows(out$psi_list)

# Summarize properly (quantile-based)
psi_otu_summary <- psi_draws %>%
    group_by(OTU) %>%
    summarise(
        psi_mean = mean(plogis(eta), na.rm = TRUE),
        psi_lwr  = quantile(plogis(eta), 0.025, na.rm = TRUE),
        psi_upr  = quantile(plogis(eta), 0.975, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    arrange(psi_mean) %>%
    mutate(OTU = factor(OTU, levels = OTU))

# Plot
ggplot(psi_otu_summary, aes(x = psi_mean, y = OTU)) +
    geom_point(color = "darkgreen", size = 2) +
    geom_errorbarh(
        aes(xmin = psi_lwr, xmax = psi_upr),
        height = 0.2,
        color = "darkgreen"
    ) +
    labs(
        title = "Caterpillar Plot: Occupancy Probability by OTU",
        x = "Occupancy Probability (ψ)",
        y = "OTU"
    ) +
    theme_minimal() +
    theme(
        axis.text.y = element_text(size = 8),
        plot.title = element_text(hjust = 0.5)
    )


```

---

## 4.2 Detection Probability Caterpillar Plot

```{r plot-detection, fig.width=10, fig.height=9}

library(dplyr)
library(ggplot2)

# Combine retained iterations
p_detect_draws <- dplyr::bind_rows(out$p_detect_list)

# Convert cloglog eta to detection probability
p_detect_draws <- p_detect_draws %>%
  mutate(
    p_detect = 1 - exp(-exp(eta))
  )

# Summarise at OTU level
p_detect_otu_summary <- p_detect_draws %>%
  group_by(OTU) %>%
  summarise(
    p_detect_mean = mean(p_detect, na.rm = TRUE),
    p_detect_lwr  = quantile(p_detect, 0.025, na.rm = TRUE),
    p_detect_upr  = quantile(p_detect, 0.975, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(p_detect_mean) %>%
  mutate(
    OTU = factor(OTU, levels = OTU)
  )

ggplot(
  p_detect_otu_summary,
  aes(x = p_detect_mean, y = OTU)
) +
  geom_point(size = 2) +
  geom_errorbarh(
    aes(
      xmin = p_detect_lwr,
      xmax = p_detect_upr
    ),
    height = 0.2
  ) +
  labs(
    title = "Caterpillar Plot: Detection Probability by OTU",
    x = "Detection probability",
    y = "OTU"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    plot.title = element_text(hjust = 0.5)
  )
```

---

## 4.3 Occupancy by treatment

```{r plot-psi-treatment, fig.width=10, fig.height=9}

library(dplyr)
library(ggplot2)

treatment_col <- "Samplingmonth"
site_col <- "site_month"
otu_col <- "OTU"

# ------------------------------------------------------------
# Combine retained occupancy draws
# ------------------------------------------------------------

psi_draws <- dplyr::bind_rows(out$psi_list)

# ------------------------------------------------------------
# Extract site-level treatment information
# ------------------------------------------------------------

treatment_info <- out$long_df %>%
  dplyr::select(
    dplyr::all_of(c(site_col, otu_col, treatment_col))
  ) %>%
  dplyr::distinct()

# ------------------------------------------------------------
# Join treatment information correctly:
# site × OTU, NOT OTU alone
# ------------------------------------------------------------

psi_draws <- psi_draws %>%
  dplyr::left_join(
    treatment_info,
    by = c(site_col, otu_col)
  )

# Check that treatment was matched
stopifnot(
  !any(is.na(psi_draws[[treatment_col]]))
)

# ------------------------------------------------------------
# Transform logit eta -> occupancy probability
# ------------------------------------------------------------

psi_draws <- psi_draws %>%
  dplyr::mutate(
    psi = stats::plogis(.data$eta)
  )

# ------------------------------------------------------------
# Summarise by OTU and treatment
# ------------------------------------------------------------

psi_data <- psi_draws %>%
  dplyr::group_by(
    .data[[otu_col]],
    .data[[treatment_col]]
  ) %>%
  dplyr::summarise(
    psi_mean   = mean(.data$psi, na.rm = TRUE),
    psi_median = stats::median(.data$psi, na.rm = TRUE),
    psi_lwr    = stats::quantile(
      .data$psi,
      0.025,
      na.rm = TRUE
    ),
    psi_upr    = stats::quantile(
      .data$psi,
      0.975,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

# ------------------------------------------------------------
# Order OTUs according to overall occupancy
# ------------------------------------------------------------

psi_data <- psi_data %>%
  dplyr::group_by(.data[[otu_col]]) %>%
  dplyr::mutate(
    otu_mean = mean(.data$psi_mean, na.rm = TRUE)
  ) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(.data$otu_mean)

otu_levels <- unique(psi_data[[otu_col]])

psi_data[[otu_col]] <- factor(
  psi_data[[otu_col]],
  levels = otu_levels
)

# ------------------------------------------------------------
# Plot
# ------------------------------------------------------------

ggplot(
  psi_data,
  aes(
    x = psi_mean,
    y = .data[[otu_col]],
    color = .data[[treatment_col]]
  )
) +
  geom_point(
    position = position_dodge(width = 0.5),
    size = 2
  ) +
  geom_errorbarh(
    aes(
      xmin = psi_lwr,
      xmax = psi_upr
    ),
    position = position_dodge(width = 0.5),
    height = 0.2
  ) +
  labs(
    title = "Occupancy Probability by OTU and Sampling Month",
    x = expression("Occupancy probability (" * psi * ")"),
    y = "OTU",
    color = "Sampling month"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    plot.title = element_text(hjust = 0.5)
  )

```

---

## 4.4 Clustered Heatmap of Occupancy Probabilities

```{r heatmap-psi, fig.width=10, fig.height=9}



library(dplyr)
library(tidyr)
library(tibble)
library(pheatmap)

site_col <- "site_month"
otu_col  <- "OTU"

# ------------------------------------------------------------
# Step 1: Combine retained occupancy draws
# ------------------------------------------------------------
psi_draws <- dplyr::bind_rows(out$psi_list)

# Check
names(psi_draws)
head(psi_draws)

# ------------------------------------------------------------
# Step 2: Convert logit eta -> occupancy probability
# ------------------------------------------------------------
psi_long <- psi_draws %>%
    dplyr::mutate(
        psi = stats::plogis(.data$eta)
    )

# ------------------------------------------------------------
# Step 3: Average across retained EM-like iterations
# for each site × OTU combination
# ------------------------------------------------------------
psi_site_otu <- psi_long %>%
    dplyr::group_by(
        .data[[site_col]],
        .data[[otu_col]]
    ) %>%
    dplyr::summarise(
        psi = mean(.data$psi, na.rm = TRUE),
        .groups = "drop"
    )

# ------------------------------------------------------------
# Step 4: Convert to site × OTU matrix
# ------------------------------------------------------------
psi_matrix <- psi_site_otu %>%
    tidyr::pivot_wider(
        names_from  = dplyr::all_of(otu_col),
        values_from = psi
    ) %>%
    tibble::column_to_rownames(site_col) %>%
    as.matrix()

# ------------------------------------------------------------
# Step 5: Inspect
# ------------------------------------------------------------
dim(psi_matrix)
psi_matrix[1:min(5, nrow(psi_matrix)),
           1:min(5, ncol(psi_matrix)),
           drop = FALSE]

# ------------------------------------------------------------
# Step 6: Heatmap
# ------------------------------------------------------------
pheatmap::pheatmap(
    psi_matrix,
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    main = "Occupancy Probability by Site and OTU",
    border_color = NA
)
```


```{r model-comparison-full, echo=FALSE}

##For example 

  models = list(
  
  A = function(ps) FitModel(
    phyloseq = ps,
    site_col = "site_month",
    sample_col = "Name",
    replicate_col = "Replicate",
    otu_col = "OTU",
    count_col = "y",
    
    occupancy_formula = z_sim ~ 1 + (1 | OTU),
    capture_formula   = a_sim ~ 1 + (1 | OTU),
    
    abundance_formula =
      y ~ offset(log(total_reads)) +
      (1 | OTU) +
      (1 | Name / OTU),
    
    abundance_family = "nbinom",
    
    min_species_sum = 10,
    min_detection_replicates = 1,
    abundance_threshold = 0,
    
    n_iter  = 10,
    burn_in = 1,
    
    verbose = TRUE,
    n_sim_ci = 50
  ),
  
  B = function(ps) FitModel(
    phyloseq = ps,
    site_col = "site_month",
    sample_col = "Name",
    replicate_col = "Replicate",
    otu_col = "OTU",
    count_col = "y",
    
    occupancy_formula = z_sim ~ 1 + (1 | OTU),
    capture_formula   = a_sim ~ 1 + (1 | OTU),
    
    abundance_formula =
      y ~ offset(log(total_reads)) +
      (1 | OTU) +
      (1 | `Sampling.area.Name` / OTU) +
      (1 | Name / OTU),
    
    abundance_family = "nbinom",
    
    min_species_sum = 10,
    min_detection_replicates = 1,
    abundance_threshold = 0,
    
    n_iter  = 10,
    burn_in = 1,
    
    verbose = TRUE,
    n_sim_ci = 50
  ),
  
  C = function(ps) FitModel(
    phyloseq = ps,
    site_col = "site_month",
    sample_col = "Name",
    replicate_col = "Replicate",
    otu_col = "OTU",
    count_col = "y",
    
    occupancy_formula = z_sim ~ 1 + (1 | OTU),
    capture_formula   = a_sim ~ 1 + (1 | OTU),
    
    abundance_formula =
      y ~ (1 | OTU) +
      (1 | siteCode / OTU) +
      (1 | Samplingmonth / OTU) +
      (1 | Name / OTU),
    
    abundance_family = "nbinom",
    
    min_species_sum = 10,
    min_detection_replicates = 1,
    abundance_threshold = 0,
    
    n_iter  = 10,
    burn_in = 1,
    
    verbose = TRUE,
    n_sim_ci = 50
  )
)


cv_rep <- compare_models_cv_multilevel(
  phyloseq = ps,
  models = models,
  site_col = "site_month",
  sample_col = "Name",
  replicate_col = "Replicate",
  otu_col = "OTU",
  count_col = "y",
  cv_level = "replicate",
  K = 3
)

cv_sample <- compare_models_cv_multilevel(phyloseq = ps,
                                          models = models,
                                          site_col = "site_month",
                                          sample_col = "Name",
                                          replicate_col = "Replicate",
                                          otu_col = "OTU",
                                          count_col = "y", cv_level = "sample",K = 5)
cv_site   <- compare_models_cv_multilevel(phyloseq = ps,
                                          models = models,
                                          site_col = "site_month",
                                          sample_col = "Name",
                                          replicate_col = "Replicate",
                                          otu_col = "OTU",
                                          count_col = "y", cv_level = "site",K = 5)

cv_otu   <- compare_models_cv_multilevel(phyloseq = ps,
                                          models = models,
                                          site_col = "site_month",
                                          sample_col = "Name",
                                          replicate_col = "Replicate",
                                          otu_col = "OTU",
                                          count_col = "y", cv_level = "OTU",K = 5)




evaluate_cv(cv_rep)
evaluate_cv(cv_sample)
evaluate_cv(cv_site)
evaluate_cv(cv_otu)

